---
title: "CLEANING_TRANSFO_HALASEN"
output: html_document
date: "2023-04-12"
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

# *0 - MEMOS - HELP*

```{r KNITR global options: TO DO NOT INCLUDE}
#| include = F
 
knitr::opts_chunk$set(echo = F, include = F)
```

controls: - alt+o = close all - shift+alt+o = open all - ctrl + shift +
c to add or delete "#" the selected lines

dput(as.character(samples_missing_postscata)) \# convert into the format
c(x; y; ...; z) as.data.frame.matrix() \# table to dataframe for ggplot:
position = fill -\> relative abundance

DURBIN WATSON TEST: - Durbin Watson statistic is a test for
autocorrelation in a regression model's output. - The DW statistic
ranges from zero to four, with a value of 2.0 indicating zero
autocorrelation. Values below 2.0 mean there is positive autocorrelation
and above 2.0 indicates negative autocorrelation - A rule of thumb is
that DW test statistic values in the range of 1.5 to 2.5 are relatively
normal significantly small p-value casts doubt on the validity of the
null hypothesis and indicates autocorrelation among residuals.

BREUSCH PAGAN TEST: - Null Hypothesis (H0): Homoscedasticity is present
(the residuals are distributed with equal variance) - Alternative
Hypothesis (HA): Heteroscedasticity is present (the residuals are not
distributed with equal variance)

FOR MODEL: -- Variables: - zone : fixed - year : fixed - stream: fixed ?
random ? - orientation: random ? not considered ? - plot: random

-- symbols: + : include - : delete : : interaction between variables \*

:   variables + their interaction \| : conditionning = X\|Y -\> include
    X given Y 1 : intercept (can be deleted)

-   Random intercept: m1 \<- lmer(Reaction \~ Days + (1 \| Subject),
    data = sleepstudy)

-   Random slope: m2 \<- lmer(Reaction \~ Days + (0 + Days\|Subject),
    data = sleepstudy)

-   Correlated random intercept and slope: m3 \<- lmer(Reaction \~
    Days + (1 + Days\|Subject), data = sleepstudy)

-   Uncorrelated random intercept and slope: m4 \<- lmer(Reaction \~
    Days + (1\|Subject) + (0 + Days\|Subject), data = sleepstudy)

<https://stats.stackexchange.com/questions/31569/questions-about-how-random-effects-are-specified-in-lmer>
for constructing model with random effects

-   to transform summary() into df summary_ecm_df \<-
    data.frame(unclass(summary(ecmtax_df)), check.names = FALSE,
    stringsAsFactors = FALSE)

# *1 - REMAINS TO DO*

-   deciding about effect of variables (model -\> stream + orientation +
    pool) -- find a way to compute samples together if no significance
    (knowing that if we pool samplings UP-LO N-S together, we need to do
    it for all of them for comparison) -- if significance from
    environmental variables : consider using them as explanatory
    variables. --- PCA and testing correlation

-testing autocorrelation in plots (if we want to analyse them
independantly)

-   venn pb (all common, nothing specific)
-   permanova/adonis
-   commenting \##
-   synthetic tab for MM

\~\~ 27/03/23

-   Automatic taxonomic attribution for the 3050 not manually IDed
    clusters. Using a local database (Soil Inventory 2012 - 2019) on
    computer and ID automatically Genus based on 90% threshold
    similarity for sequences. Use the closest match if this is \>90%,
    otherwise =\> unknown Then taxa-ecologic attribution with
    "FUNGuild-like" database directly into R. \# You don't have to use
    FUNGuild -- species are already denoted as "ECM" in the Soil
    Inventory database (but if you use FUNGuild you should hopefully get
    the same output.

-   If sequencing depth of samples are even: -- Rarefying OTU matrix by
    considering all 4050 OTU clusters. Apply rarefaction for the lowest
    number of reads (ie sequencing depth) for each sample. \# Yes, and
    if you have some few samples with few reads, you may consider
    omitting these samples rather than rarefying all samples to a very
    low sequencing depth. Alternatively, you can skip rarefying and
    instead include sqrt(sequencing depth) as covariate in statistical
    models (but it is a bit more complicated to explain and interpret).

    -\> But what about the really low number of reads in for some
    clusters among samples (especially clearcut where available DNA is
    low) ?

-   If not: -- Doing model and : sqrt(reads) \# exactly, however fungal
    DNA is usually not so low on clearcuts (although it is mainly
    free-living fungi and not ECM).

-   Biais in FUNGuild to consider: "ECM" are far more studied and then
    so, when there is a matching attribution -\> confident ; but for "SF
    (Saprophytic Fungi)", less sure. A lot of them are not identified as
    SF in database, even if they are, because of lacking studies. -\>
    ECM and "others" (saprobes, root-associated, parasitic)

-   Halasen edge study: spatial dependence among samples inside a same
    plot (distant from 7.5 meters from each other) but independence of
    others plots -\> Should we consider the factor "plot" as random
    variable or just the concerned ones ? If so, how to do it ? GLMM,
    paidarkred-ANOVA ? \# Yes, you should probably use "plot" as a
    random variable in a GLMM.

-   to limit effect of dominant OTU (high nb of reads): Hellinger
    transformation is sqrt (reads of species / total reads in sample) ie
    sqrt (relative abudances)

-   Zero oSCurence of OTUs or no reads in some samples -\> leads to a
    biais: -- considering the samples with 0 OTU counted: min richness =
    0 & N(samples) unchanged -\> good for design

-- not considering them (ie erasing samples with 0 from dataset): min
richness = 1 & N(samples) darkreduced -\> no significant differences in
mean comparison tests"

Rarefaction Data.rar\<-as.data.frame(rrarefy(data, min(rowSums(data))))

Hellinger transformation

data.hel\<-as.data.frame(t(apply(data,1,function(x) sqrt(x/sum(x)))))

For automatic ID you will have to make a local blast database
(<https://www.ncbi.nlm.nih.gov/books/NBK569841/>). I do this on the
Mykopat grid. Then you blast your sequences against this database, using
a custom output format. This is what a script can look like if you run
it on the Mykopat grid:

\$-l h_rt=24:0:0 \$-l h_vmem=1G \$-m ae module load ncbi-blast+ blastn
--db local_databse -query query_sequences.fst -out Output_file.txt
-remote -max_target_seqs 1 -outfmt "6 qseqid sseqid evalue pident"

# *2 - LIBRARY - OBJECTS for later*

```{r packages}
library(devtools) # for install_github etc. 
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(dplyr) # A Grammar of Data Manipulation
library(tidyverse) # A Grammar of Data Manipulation
library(vegan) # ecological statistical analysis
library(phyloseq) # Handling and analysis of high-throughput microbiome census data 
library(forcats) # Tools for Working with Categorical Variables
library(MiscMetabar) # Analyze and visualize metabarcoding data
library(circlize) # circular visualization
library(ggVennDiagram) # venn diagrams 
library(phyloseq.extended) # phyloseq packages with more options
library (microbiomeutilities) # for rarefaction curves
library(multcompView) # Visualizations of Paired Comparisons
library(plyr) # tools for cleaning 
library(pals) # color palette
library(ade4) # Analysis Data Ecological Environmental in the framework of Euclidean Exploratory methods
library(lme4) # Linear, generalized linear, and nonlinear mixed models
library(lmtest) # Diagnostic Checking in Regression Relationships
library(reshape2)
library(data.table)
library(viridis) # palettes and themes
library(hrbrthemes) # palettes and themes
library(emmeans) # tukey test for GLM
library(ISLR) # McFadden’s R-Squared for GLM
library(ape) # Analyses of Phylogenetics and Evolution
library(sjPlot) # for glmm visualization
# undersampling lib
library(purrr) # purrr: Functional Programming Tools
library(randomForest) # randomForest implements Breiman's random forest algorithm (based on Breiman and Cutler's original Fortran code) for classification and regression
library(caret)
library(phylosmith) # A supplementary package to build on the phyloseq package
library(brms) # Bayesian Regression Models using 'Stan'
library(glmmTMB) # Fit a generalized linear mixed model (GLMM) using Template Model Builder (TMB).
library(gtsummary)
library(pairwiseAdonis) # pari-wise analysis for PermANOVAs 
library(gplots)
library(zoom)
library(recolorize)
#library(FUNGuildR)
```

```{r color palette}
pals::pal.bands(alphabet, alphabet2, cols25, glasbey, kelly, polychrome, 
  stepped, tol, watlington,
  show.names=FALSE)

funky <- grDevices::colorRampPalette(c("#A6CEE3","#1F78B4","#B2DF8A",
"#33A02C","#FB9A99","#E31A1C",
"#FDBF6F","#FF7F00","#CAB2D6",
"#6A3D9A","#FFFF99","#B15928"))
```

```{r functions}
outersect <- function(x, y) {
  sort(c(setdiff(x, y),
         setdiff(y, x)))
} # extract uncommon values between x - y 

stde <- function (x) {sqrt(var(x))/length(x)}

give.n <- function(x){
   return(c(y = mean(x), label = length(x)))
}
```

```{r dummy calc}
# hellinger : sum of reads per sample = 1 ? (as same as relative abundance ?)

#abund1 <- c(sample(0:250, n = 10))

# function if()
x <- 0
if (x < 0) {
  print("Negative number")
} else if (x > 0) {
  print("Positive number")
} else {
  print("Zero")
}

round_df <- function(x, digits) {
    # round all numeric variables
    # x: data frame 
    # digits: number of digits to round
    numeric_columns <- sapply(x, mode) == 'numeric'
    x[numeric_columns] <-  round(x[numeric_columns], digits)
    x
}

# if() is used for a single value
# to check for a vector: use ifelse instead
# can use ifelse(test, yes, ifelse(test, yes, no)) nested 

# Function to convert RGB values to HEX
rgb_to_hex <- function(red, green, blue) {
  rgb_value <- rgb(red, green, blue)
  hex_value <- col2rgb(rgb_value)
  hex_value <- sprintf("#%02X%02X%02X", hex_value[1, 1], hex_value[2, 1], hex_value[3, 1])
  return(hex_value)
}
```

```{r objects}
coltreatment <- c("CCFM" = "#FFDC9A",
                  "CUT" = "#FF989F",
                  "FOR" = "#AEDBAF") # palette for treatment

# increasing saturation for the multivariate and rarecurve plotting
colors <- c("#FFDC9A","#FF989F", "#AEDBAF")
colors <- t(col2rgb(colors)/ 255 )
colors_hex <- apply(colors, 1, function(row) rgb_to_hex(row[1], row[2], row[3]))
coltreatment_saturated <- recolorize::adjust_color(rgb_color = colors, which_colors = "all", saturation = 1.8, brightness = 1, plotting = T) # increasing saturation of all
coltreatment_saturated <- recolorize::adjust_color(rgb_color = coltreatment_saturated, which_colors = c(1,3), saturation = 1.8, brightness = 1, plotting = T) # increasing specifically forest

# Apply the function to each row of the matrix
coltreatment_saturated <- apply(coltreatment_saturated, 1, function(row) rgb_to_hex(row[1], row[2], row[3]))
coltreatment_saturated
#

# alpha diversity custom order
treat_order <- c("FOR", "CCFM", "CUT")
treat_color <- c("FOR" = "#AEDBAF", "CCFM" = "#FFDC9A", "CUT" = "#FF989F")

order_level <- c("FOR", "CCFM", "CUT") # ordination of factors (aes = factor(x, levels = order_level))
order_level_mg <- c("2012FOR", "2012CCFM", "2012CUT", "2018FOR", "2018CCMF", "2018CUT")
order_level2 <- c("FOR_12", "CCFM_12", "CUT_12", "FIR_18", "CCFM_18", "CUT_18")

# custom theme
custom_theme <- theme_minimal() + 
  theme(
    text = element_text(family = "Arial", size = 12),
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9),
    legend.background = element_rect(color = "grey"),
    panel.border = element_rect(color = "grey", fill = NA, linewidth = 0.5),
    axis.line = element_line(color = "grey", linewidth = 0.5),
  )

# setting custom theme as the default
theme_set(custom_theme)

### FOR HEATMAP CONSTRUCTION
# common breaks 
com_breaks <- c(0, 5, 15, 30, 50, 70)

# for annotation columns 
zone <- c("FOR", "CCFM", "CUT", "FOR", "CCFM", "CUT")
year <- c("2012", "2012", "2012", "2018", "2018", "2018")
df_color <- data.frame(zone, year)
rownames(df_color) <- rownames(occurence_fanga)
# opposite nested:
zone_bis <- c("FOR", "FOR", "CCFM", "CCFM", "CUT", "CUT")
year_bis <- c("2012", "2018", "2012", "2018", "2012", "2018")
df_color_bis <- data.frame(zone_bis, year_bis)
rownames(df_color_bis) <- paste(df_color_bis$zone_bis, df_color_bis$year_bis, sep = "_")

# for annotation colors
anot_colors_year <- list(year = c("2012" = "ivory2", "2018" = "ivory4"))
anot_colors_treat <- list(zone = c("FOR" = "#AEDBAF", "CCFM" = "#FFDC9A", "CUT" = "#FF989F"))
```

# *3 - DATA IMPORTATION*

```{r raw_data}
setwd("C:/Users/Utilisateur/OneDrive/STAGE-2023_SUEDE/cutting_ecm-main_branch/TRANSF_DATA2") 
# SLU: H:/REDACTION/RESULTS/cutting_ecm/TRANSF_DATA2
# HOME: C:/Users/Utilisateur/OneDrive/STAGE-2023_SUEDE/cutting_ecm-main_branch/TRANSF_DATA2
#data run

# fruitbody dataset
fruit <- read.table("fruit.txt", header = T, sep="\t",na.strings=T,dec=".") 
rownames(fruit) <- fruit$X # putting X column into rownames
fruit <- fruit[, -1] # supressing obsolete column
colnames(fruit) <- gsub("X", "", colnames(fruit)) # supressing "X" from colnames 

# sequencing run metadata dataset
run_data <- read.table("run_data.txt", header=T, sep="\t",na.strings=T,dec=".")
colnames(run_data)

# all clusters
raw_clust <- read.table("raw_clusters.txt",header=T,sep="\t",na.strings=T,dec=".") %>%
  arrange(desc(Cluster_Size)) # importing and arrange by decreasing cluster size
raw_clust[raw_clust == ""] <- NA # putting NA when empty
raw_clust[, 4] <- as.character(raw_clust[, 4]) # reference as.character
raw_clust$Cluster_ID <- gsub('scata5670_', 'c', raw_clust$Cluster_ID) # simplifying clusters names
rownames(raw_clust) <- raw_clust$Cluster_ID # attributing each row to their cluster_id (OTU name) 

order <- raw_clust$Cluster_ID # creating a numerical vector order (by deleting the "c"), so we can arrange later by number
order <- gsub('c','', raw_clust$Cluster_ID) 
raw_clust <- data.frame(cbind(order, raw_clust)) # binding string and df
raw_clust[, "order"] <- as.numeric(raw_clust[, "order"]) # transf. in num. variable

# subsetting fungi X non-fungi
raw_f_clust <- subset(raw_clust, !grepl("PLANT|NF", raw_clust[, 5]))
raw_nf_clust <- subset(raw_clust, grepl("PLANT|NF", raw_clust[, 5]))

# taxonomic and ecological traits table
# taxeco = only the 1st 1000 clusters manually verified / identified when needed. 
taxeco_id <- read.table("tax_table.txt",header=T,sep="\t") # change #N/A in excel in NA
length(taxeco_id) # number of columns 
length(rownames(taxeco_id)) # number of rows (ie clusters, 1st 1000 fungi)
taxeco_id[, c(3:length(taxeco_id))] <- lapply(taxeco_id[, c(3:length(taxeco_id))], as.factor) # transform character strings into factors
rownames(taxeco_id) <- taxeco_id$Cluster_ID # rownames of corresponding clusters
```

```{r test funguild (to develop)}
attrib_raw <- funguild_assign(raw_clust, tax_col = "Genus")
# could be interesting 
```

# *4 - DATA MANIPULATION*

```{r tax-eco table manipulation }
# creating the variable "ID_TYPE" (auto or manual) for the OTU identification. 
## for this, we check the match between the OTU_ID raw_f_clust (ie the whole dataset) and taxeco (ie manually reviewed table)
id_tax <- raw_f_clust %>%
  filter(Cluster_ID %in% taxeco_id$Cluster_ID) %>%
  arrange(order) # filtering taxtable identified manually
id_tax <- cbind("MANUAL_ID", id_tax)

unid_tax <- raw_f_clust %>%
  filter(!Cluster_ID %in% taxeco_id$Cluster_ID) %>%
  arrange(order) # filtering taxtable not identified 

tax_f <- dplyr::bind_rows(id_tax, unid_tax) # merging "id_tax" and "unid_tax" together
colnames(tax_f)[1] <- "ID_TYPE"
tax_f$ID_TYPE[is.na(tax_f$ID_TYPE)] <- "AUTO_ID" # replacing the "ID_TYPE" column with new value

taxeco_id <- cbind("MANUAL_ID", taxeco_id) 
colnames(taxeco_id)[1] <- "ID_TYPE"

# id_tax and taxeco_id should be identical 
match_id <- taxeco_id$Cluster_ID %in% id_tax$Cluster_ID
as.factor(match_id) %>% 
  levels()
which(match_id == "FALSE", arr.ind = T) # mismatch at row 240
## not same length row between id_tax (998) and taxeco (999): not normal, can cause offset
## can be explained by forgetting an OTU during manual process that lead to offset everything.

# cluster is identified as "Fungi": no relevance in our analysis, we can then delete it
taxeco_id <- taxeco_id[-240,] # deletion of row 240

match_id <- taxeco_id$Cluster_ID %in% id_tax$Cluster_ID # checking if resolved
as.factor(match_id) %>% 
  levels()
which(match_id == "FALSE", arr.ind = T) # good. no more mismatch

## We need to merge the "tax_f" which includes whole fungal community but is lacking tax-ecological traits (only attributed from 1st 1000 OTUs), with "taxeco_id" that includes their reference from SCATA with the dataset manually IDed + FunGuild attribution. 
taxeco_id$Cluster_Size <- id_tax$Cluster_Size # reattributing the proper cluster_size values

taxeco2 <- merge(tax_f, taxeco_id, all = T) %>%
  arrange(order) # merging taxeco and idtaxeco  

rownames(taxeco2) <- taxeco2$Cluster_ID

# attribution of automatic ID (ie reference) for OTU that were not manually reviewed
which(taxeco2[,"ID_TYPE"] == "MANUAL_ID") %>% length() # finding the rows NOT concerned
taxeco2$OTU_ID <- as.character(taxeco2$OTU_ID) # putting as character for next line
taxeco2$OTU_ID <- c(taxeco2$OTU_ID[1:998],  taxeco2$Reference[999:length(taxeco2$Reference)]) # merging reference and otu_id together
# AUTO_ID OTUs have tax-ecological traits column from FunGuild with NAs values but have now value in OTU_ID column
  
# subsetting interesting variables by keeping only tax-ecological traits 
colnames(taxeco2)
taxeco <- taxeco2[, c(17:24,1)] 
taxeco <- sapply(taxeco, as.character) 
rownames(taxeco) <- rownames(taxeco2) # cluster name in row
colnames(taxeco)
colnames(taxeco)[8] <- "Fruitbody" # change variable name
```

```{r sample data}
# sample data
sam <- read.table("sample_table.txt",header=T,sep="\t",na.strings = T) 
sam_no_match <- read.table("samples_no_match.txt",header=T,sep="\t") 

sam[sam==""] <- NA # filling the blanck by NAs
rownames(sam) <- sam$sample_id # after transposing, resolving pb with headers

sam_all <- filter(sam, site %in% c("Halasen", "Fangamon", "Halasen_p_5"))
sam <- sam_all
```

```{r changing levels of factor for sam}
# for Halasen CUT 
factor(sam$specification) %>% levels()
sam$zone <- plyr::revalue(sam$specification,
                           c("edge_north"="edge", 
                             "edge_south"="edge",
                             "kant_norra"="edge", 
                             "middle_lower"="middle", 
                             "middle_upper"="middle",
                             "midle"="middle"))
as.factor(sam$zone) %>% droplevels() %>% levels()

sam$specification_value <- plyr::revalue(sam$specification,
                           c("edge_north"="N", 
                             "edge_south"="S",
                             "kant_norra"="N", 
                             "middle_lower"="LO", 
                             "middle_upper"="UP",
                             "midle"= NA,
                             "middle"= NA))

sam <- within(sam, zone[zone=="middle" & treatment=="forest"] <- "FOR") 

sam$zone <- plyr::revalue(sam$zone,
                           c("middle"="CUT", 
                             "edge"="CCFM"))
as.factor(sam$zone) %>% droplevels() %>% levels()

# for Halasen thinning + Fangamon
sam <- within(sam, zone[site == "Halasen_p_5" & treatment=="control"] <- "FOR") 
sam <- within(sam, zone[site == "Halasen_p_5" & treatment=="clearcut"] <- "CUT") 
sam <- within(sam, zone[site == "Halasen_p_5" & treatment=="p_5"] <- "CCFM")
sam <- within(sam, zone[site == "Fangamon" & treatment=="forest_control"] <- "FOR") 
sam <- within(sam, zone[site == "Fangamon" & treatment=="clearcut"] <- "CUT") 
# merging burn - not burned for fangamon (cf RESULTS-MODEL Fangamon)
sam <- within(sam, zone[site == "Fangamon" & treatment=="shelter_without_burn"] <- "CCFM") # change with CCFM_NB" to test model significance
sam <- within(sam, zone[site == "Fangamon" & treatment=="shelter_burn"] <- "CCFM") # change with CCFM_B" to test model significance

# deleting root_seedling samples
sam <- subset(sam, !grepl("seedling_roots", sam$specification))

### subsetting

# subsetting design samples per site 
sam_hala <- subset(sam, sam$site == "Halasen") # 61 samples # 64 initially!
# missing 2 CUT + 1 FOR
sam_hala5p <- subset(sam, sam$site == "Halasen_p_5")
sam_fanga <- subset(sam, sam$site == "Fangamon")
sam_controls_na <- filter(sam, !site %in% c("Halasen", "Fangamon", "Halasen_p_5")) # 19 samples (or controls)

## tables
table(sam$zone, sam$year, sam$site)
```

```{r random resampling -> balanced design}
# Set the seed for reproducibility
set.seed(123)
# ## Fangamon -> merging CCFM experiments and resampling 2018
# table(sam_fanga$zone, sam_fanga$year)
 for_subset_fanga12 <- sam_fanga[sam_fanga$zone == "CCFM" & sam_fanga$year == "2012", ]
 for_subset_fanga12 <- for_subset_fanga12[sample(nrow(for_subset_fanga12), 9), ]

sam_fanga_filtered <- subset(sam_fanga, !(sam_fanga$year == "2012" & sam_fanga$zone == "CCFM"))
sam_fanga <- rbind(for_subset_fanga12, sam_fanga_filtered)
table(sam_fanga$zone, sam_fanga$year)
# 
# ## Halasen FOR 2012
# # Subsample 8 observations from the "FOR" category
# for_subset_hala12 <- sam_hala12[sam_hala12$zone == "FOR", ]
# for_subset_hala12 <- for_subset_hala12[sample(nrow(for_subset_hala12), 8), ]
# 
# # Combine the subsampled "FOR" category with the other categories
# equal_sample_hala12 <- rbind(
#   sam_hala12[sam_hala12$zone == "CUT", ],
#   sam_hala12[sam_hala12$zone == "CCFM", ],
#   for_subset_hala12
# )
# ## need to bind this new halasen 2012 table with the rest
# 
# ## Halasen FOR 2018
# # Subsample 8 observations from the "FOR" category
# for_subset_hala18 <- sam_hala18[sam_hala18$zone == "FOR", ]
# for_subset_hala18 <- for_subset_hala18[sample(nrow(for_subset_hala18), 8), ]
# 
# # Combine the subsampled "FOR" category with the other categories
# equal_sample_hala18 <- rbind(
#   sam_hala18[sam_hala18$zone == "CUT", ],
#   sam_hala18[sam_hala18$zone == "CCFM", ],
#   for_subset_hala18
# )
# 
# # Generate the table of sample sizes for each category
# sam_hala12 <- equal_sample_hala12
# sam_hala18 <- equal_sample_hala18
# 
# sam_hala1218 <- rbind(sam_hala12, sam_hala18) # do the subsampling after physeq creation 
# sam_hala <- sam_hala1218
# table(sam_hala$zone, sam_hala$year)
# 
# 
# ## merging all together
# sam <- merge(merge(sam_fanga, sam_hala, all = TRUE), sam_hala5p, all = TRUE)
# rownames(sam) <- sam$sample_id
# 
# table(sam$zone, sam$year, sam$site)
```

```{r otu table}
otu_all <- read.table("otu_table.txt",header=F,sep="\t",na.strings = T) 
otu_all = setNames(data.frame(t(otu_all[,-1])), otu_all[,1]) # transpose + 1st column (ie samples set as header
## samples = columns, otu abundance = rows
colnames(otu_all)[1] <- "Cluster_ID" # after transposing, resolving pb with headers
nrow(otu_all) # total number of all clusters (ie fungi ecm or no-ecm, plants)
rownames(otu_all) <- otu_all$Cluster_ID
otu_all <- otu_all[-1] # deleting 1st column (cluster_id is now in rownames)
colnames(otu_all)
cluster_names_otu_all <- rownames(otu_all)

# removing clusters with 0 rowsums 
## relics of PCR mismatch tag (still amplified)
## but clusters still there after the exclusion of mismatch tags (eg ITS7-49 x ITS4-48)
otu_all <- sapply(otu_all, as.numeric) %>% as.data.frame()
rownames(otu_all) <- cluster_names_otu_all
which(rowSums(otu_all) == 0) %>% length() # clust with 0 reads = 557 (4050 - 557 = 3493)
otu <- otu_all %>%
  filter(!rowSums(otu_all) ==0) # suppression of these 557 clusters from otu matrix
nrow(otu)
ncol(otu)

# controls
not_samples_kept <- c("sITS-standard2", "sITS-standard1", "sPCR_neg_1", "sPCR_neg_2", "sQ2") # cf MM_RESULTS.Rmd ```{r DESIGN SAMPLES}  to understand how they are found 

samples_kept_pq <- otu[, colnames(otu) %in% not_samples_kept] # table of control samples and clusters in row. we want to identity and deleted clusters with reads in these. (potential contaminations)

clusters_in_controls <- which(samples_kept_pq > 0, arr.ind = T) %>% 
rownames() %>%  as.factor() %>% levels() %>% print() # 12 clusters found in controls samples
otu_samples_out <- otu[!(rownames(otu) %in% not_samples_kept),] # suppressing these samples from otu matrix
ncol(otu_samples_out)
nrow(otu_samples_out) 

# excluding NF clusters to rarefy only with fungi
otu_fungi_all <- otu_samples_out %>%
  filter(rownames(otu_samples_out) %in% rownames(taxeco)) # taxeco has only fungi clusters
ncol(otu_fungi_all)
nrow(otu_fungi_all)

# creating otu_fungi_hala only for hala to rarefy on the unique same site
otu_fungi_t0 <- t(otu_fungi_all) %>% as.data.frame()

# fangamon
otu_fungi_fanga <- otu_fungi_t0 %>%
  filter(rownames(otu_fungi_t0) %in% rownames(sam_fanga)) 
nrow(otu_fungi_fanga)
nrow(sam_fanga)
# halasen CUT
otu_fungi_hala <- otu_fungi_t0 %>%
  filter(rownames(otu_fungi_t0) %in% rownames(sam_hala)) 
nrow(otu_fungi_hala)
nrow(sam_hala) # 3 
# halasen 5p
otu_fungi_hala5p <- otu_fungi_t0 %>%
  filter(rownames(otu_fungi_t0) %in% rownames(sam_hala5p)) 
nrow(otu_fungi_hala5p)
nrow(sam_hala5p) # 2
```

```{r otu <1% suppresion all}
# transpo
otu_fungi_hala <- t(otu_fungi_hala)
otu_fungi_hala5p <- t(otu_fungi_hala5p)
otu_fungi_fanga <- t(otu_fungi_fanga)

# transfo into matrix
otu1 <- matrix(as.numeric(unlist(otu_fungi_hala)),
              ncol = ncol(otu_fungi_hala),
              nrow = nrow(otu_fungi_hala)) 
rownames(otu1) <- rownames(otu_fungi_hala)
colnames(otu1) <- colnames(otu_fungi_hala)

otu2 <- matrix(as.numeric(unlist(otu_fungi_hala5p)),
              ncol = ncol(otu_fungi_hala5p),
              nrow = nrow(otu_fungi_hala5p)) 
rownames(otu2) <- rownames(otu_fungi_hala5p)
colnames(otu2) <- colnames(otu_fungi_hala5p)

otu3 <- matrix(as.numeric(unlist(otu_fungi_fanga)),
              ncol = ncol(otu_fungi_fanga),
              nrow = nrow(otu_fungi_fanga)) 
rownames(otu3) <- rownames(otu_fungi_fanga)
colnames(otu3) <- colnames(otu_fungi_fanga)

# suppression of <1 %
## creating a new matrix where nb. of reads are transformed in relative abundance / sample to determine which clusters have <1% of relative abundance. 
otu_relative1 <- prop.table(otu1, margin = 2) # margin = 2 -> column sum and relative abundances 
otu_001_transf1 <- ifelse(otu_relative1 > 0.01,otu1,0) # creating a new matrix where <1% reads are transformed into 0
otu_001_1 <- otu_001_transf1

otu_relative2 <- prop.table(otu2, margin = 2) 
otu_001_transf2 <- ifelse(otu_relative2 > 0.01, otu2, 0) 
otu_001_2 <- otu_001_transf2

otu_relative3 <- prop.table(otu3, margin = 2) 
otu_001_transf3 <- ifelse(otu_relative3 > 0.01, otu3, 0) 
otu_001_3 <- otu_001_transf3
```

```{r otu rarefaction all}
# rarefaction based on all 
otu_t01 <- t(otu_001_1)
otu_t1 <- matrix(as.numeric(unlist(otu_t01)),
              ncol = ncol(otu_t01),
              nrow = nrow(otu_t01))
rownames(otu_t1) <- rownames(otu_t01)
colnames(otu_t1) <- colnames(otu_t01) 

otu_t02 <- t(otu_001_2)
otu_t2 <- matrix(as.numeric(unlist(otu_t02)),
              ncol = ncol(otu_t02),
              nrow = nrow(otu_t02))
rownames(otu_t2) <- rownames(otu_t02)
colnames(otu_t2) <- colnames(otu_t02)

otu_t03 <- t(otu_001_3)
otu_t3 <- matrix(as.numeric(unlist(otu_t03)),
              ncol = ncol(otu_t03),
              nrow = nrow(otu_t03))
rownames(otu_t3) <- rownames(otu_t03)
colnames(otu_t3) <- colnames(otu_t03)

otu_rarefied1 <- as.data.frame(rrarefy(otu_t1, min(rowSums(otu_t1))))
otu_rarefied2 <- as.data.frame(rrarefy(otu_t2, min(rowSums(otu_t2)))) 
otu_rarefied3 <- as.data.frame(rrarefy(otu_t3, min(rowSums(otu_t3)))) 

otu_rarefied1 <- t(otu_rarefied1)
otu_rarefied1 <- matrix(as.numeric(unlist(otu_rarefied1)),
              ncol = ncol(otu_rarefied1),
              nrow = nrow(otu_rarefied1))
rownames(otu_rarefied1) <- rownames(otu_001_1)
colnames(otu_rarefied1) <- colnames(otu_001_1)

otu_rarefied2 <- t(otu_rarefied2)
otu_rarefied2 <- matrix(as.numeric(unlist(otu_rarefied2)),
              ncol = ncol(otu_rarefied2),
              nrow = nrow(otu_rarefied2))
rownames(otu_rarefied2) <- rownames(otu_001_2)
colnames(otu_rarefied2) <- colnames(otu_001_2)

otu_rarefied3 <- t(otu_rarefied3)
otu_rarefied3 <- matrix(as.numeric(unlist(otu_rarefied3)),
              ncol = ncol(otu_rarefied3),
              nrow = nrow(otu_rarefied3))
rownames(otu_rarefied3) <- rownames(otu_001_3)
colnames(otu_rarefied3) <- colnames(otu_001_3)

# merging together
otu_rarefied <- cbind(otu_rarefied1, otu_rarefied2, otu_rarefied3)

# data properties BEFORE - AFTER rarefaction
hist(colSums(otu_t)) # before raref
hist(colSums(otu_rarefied)) # after raref
#
summary(colSums(otu_001))
rd_sps_raref <- data.frame(colnames(otu_001), colSums(otu_001), colSums(otu_rarefied)) # read sum per sample
colnames(rd_sps_raref) <- c("sample_id", "sumreads_notrarefied", "sumreads_otu_rarefied")
summary(colSums(otu_001))
summary(colSums(otu_rarefied))
str(rd_sps_raref)

rd_sps_raref_gg <- reshape2::melt(rd_sps_raref[,2:3])
ggplot(rd_sps_raref_gg, aes(x = value, color = variable, fill = variable))+
  geom_histogram(aes(y = ..count..), alpha = 0.4, position = "identity")+
  geom_density(alpha = 0.3)+
  scale_y_continuous(name = "nb of samples")+
  scale_x_continuous(name = "read sum per sample", breaks = seq(0,3000 ,by = 200))+
  ggtitle("Number of samples per read sum") # raref worked good
```

```{r physeq objects}
ncol(otu_rarefied)
nrow(otu_rarefied)
nrow(sam) # 5 samples (3 hala + 2 hala5p missing) 

# delete the missing samples 
sam <- sam %>% filter(rownames(sam) %in% colnames(otu_rarefied))

# creating the physeq objects 
otu_phy_rarefied <- otu_table(otu_rarefied, taxa_are_rows = TRUE)
otu_phy_notrarefied <- otu_table(otu_notrarefied, taxa_are_rows = TRUE)
otu_phy_hellinger <- otu_table(otu_hellinger, taxa_are_rows = TRUE)
otu_phy_relative <- otu_table(otu_relative, taxa_are_rows = TRUE)
otu_phy_001kept <- otu_table(otu1, taxa_are_rows = TRUE)
taxeco_phy <- phyloseq::tax_table(taxeco)
sam_phy <- sample_data(sam)

# general physeq object
physeq <- phyloseq(otu_phy_rarefied,taxeco_phy,sam_phy) %>% 
clean_pq(remove_empty_samples = FALSE,remove_empty_taxa = T, clean_samples_names = FALSE)
physeq_rarefied <- physeq
physeq_notrarefied <- phyloseq(otu_phy_notrarefied,taxeco_phy,sam_phy) %>% clean_pq(remove_empty_samples = T,remove_empty_taxa = T, clean_samples_names = T)
# 2822 taxa removed from the <1% relative abundance suppresion. 

physeq_hellinger <- phyloseq(otu_phy_hellinger,taxeco_phy,sam_phy) %>% clean_pq(remove_empty_samples = T,remove_empty_taxa = T, clean_samples_names = T)
physeq_relative <- phyloseq(otu_phy_relative,taxeco_phy,sam_phy) %>% clean_pq(remove_empty_samples = T,remove_empty_taxa = T, clean_samples_names = T)
physeq_001kept <- phyloseq(otu_phy_001kept,taxeco_phy,sam_phy) 
# 10 taxa = erasing taxa from the control samples (among 13, 3 of them are probably also in others informative samples)
clusters_in_controls # controls to understand

# verif sample N
sample_names(physeq)
length(sample_names(physeq)) # 154 samples, as expected 
```

```{r taxa attribution mistakes}
# reattribution of correct names
physeq@tax_table[physeq@tax_table == "Gautieria_"] <- "Gautieria"
physeq@tax_table[physeq@tax_table == "Piloderma__"] <- "Piloderma"
physeq@tax_table[physeq@tax_table == "Sebacina__"] <- "Sebacina"
physeq@tax_table[physeq@tax_table == "Inocybe_geophylla"] <- "Inocybe_geophylla_(coll.)"

# changing variable levels
levels(as.factor(physeq@tax_table[,7])) # primary lifestyle levels

physeq@tax_table[,7] <- plyr::revalue(physeq@tax_table[,7],
                                      c("algal_parasite" = "other_lifestyle",
                                        "animal_parasite" = "other_lifestyle",
                                        "dung_saprotroph" = "saprotroph",
                                        "ectomycorrhizal" = "ectomycorrhizal",
                                        "lichen_parasite" = "other_lifestyle",
                                        "lichenized" = "other_lifestyle",
                                        "litter_saprotroph" = "saprotroph",
                                        "mycoparasite" = "other_lifestyle",
                                        "plant_pathogen" = "other_lifestyle",
                                        "root_endophyte" = "other_lifestyle",
                                        "soil_saprotroph" = "saprotroph",
                                        "sooty_mold" = "other_lifestyle",
                                        "unspecified_saprotroph" = "saprotroph",
                                        "wood_saprotroph" = "saprotroph"))
```

```{r subsets}
# subsetting based on sites and guilds 
# CTRL F + remove_empty_samples = T <-> remove_empty_samples = F

# guilds of whole dataset
ecm_phy <- subset_taxa(physeq, primary_lifestyle == "ectomycorrhizal") %>% 
clean_pq(remove_empty_samples = FALSE, remove_empty_taxa = T, clean_samples_names = FALSE)
sapr_phy <- subset_taxa(physeq, primary_lifestyle == "saprotroph") %>% 
clean_pq(remove_empty_samples = F,remove_empty_taxa = T, clean_samples_names = F)
other_phy <- subset_taxa(physeq, (primary_lifestyle == "other_lifestyle" | is.na(primary_lifestyle)))  %>% clean_pq(remove_empty_samples = F,remove_empty_taxa = T, clean_samples_names = F)

# by site
ecm_fanga <- subset_samples(ecm_phy, ecm_phy@sam_data$site=="Fangamon") %>% clean_pq(remove_empty_samples = FALSE,remove_empty_taxa = T, clean_samples_names = FALSE)
ecm_hala <- subset_samples(ecm_phy, ecm_phy@sam_data$site=="Halasen") %>% clean_pq(remove_empty_samples = FALSE,remove_empty_taxa = T, clean_samples_names = FALSE)
ecm_hala5p <- subset_samples(ecm_phy, ecm_phy@sam_data$site=="Halasen_p_5") %>% clean_pq(remove_empty_samples = FALSE,remove_empty_taxa = T, clean_samples_names = FALSE)

# by year
ecm_fanga12 <- subset_samples(ecm_fanga, ecm_fanga@sam_data$year == "2012") %>% clean_pq(remove_empty_samples = FALSE,remove_empty_taxa = T, clean_samples_names = FALSE)
ecm_fanga18 <- subset_samples(ecm_fanga, ecm_fanga@sam_data$year == "2018") %>% clean_pq(remove_empty_samples = FALSE,remove_empty_taxa = T, clean_samples_names = FALSE)

ecm_hala12 <- subset_samples(ecm_hala, ecm_hala@sam_data$year == "2012") %>% clean_pq(remove_empty_samples = FALSE,remove_empty_taxa = T, clean_samples_names = FALSE)
ecm_hala18 <- subset_samples(ecm_hala, ecm_hala@sam_data$year == "2018") %>% clean_pq(remove_empty_samples = FALSE,remove_empty_taxa = T, clean_samples_names = FALSE)

### suppression zero samples
ecm_zero0 <- names(which(colSums(ecm_phy@otu_table) == 0))
ecm_zero <- subset_samples(ecm_phy, !(sample_id %in% ecm_zero0))
ecm_zero@sam_data[, c(3:11)] <- lapply(ecm_zero@sam_data[, c(3:11)], as.factor) 
ecm_zero <- as_binary_otu_table(ecm_zero)

ecm_zero_fanga <- subset_samples(ecm_zero, ecm_zero@sam_data$site=="Fangamon") %>% clean_pq(remove_empty_samples = FALSE,remove_empty_taxa = T, clean_samples_names = FALSE)
ecm_zero_fanga <- merge_group(ecm_zero_fanga, "yr_treat")

ecm_zero_hala <- subset_samples(ecm_zero, ecm_zero@sam_data$site=="Halasen") %>% clean_pq(remove_empty_samples = FALSE,remove_empty_taxa = T, clean_samples_names = FALSE)
ecm_zero_hala <- merge_group(ecm_zero_hala, "yr_treat")

ecm_zero_hala5p <- subset_samples(ecm_zero, ecm_zero@sam_data$site=="Halasen_p_5") %>% clean_pq(remove_empty_samples = FALSE,remove_empty_taxa = T, clean_samples_names = FALSE)
ecm_zero_hala5p <- merge_group(ecm_zero_hala5p, "yr_treat")

fanga12_zero <- names(which(colSums(ecm_fanga12@otu_table) == 0))
fanga18_zero <- names(which(colSums(ecm_fanga18@otu_table) == 0))
# delete these zero samples to perform mds
ecm_fanga12_new <- subset_samples(ecm_fanga12, !(sample_id %in% fanga12_zero))
ecm_fanga18_new <- subset_samples(ecm_fanga18, !(sample_id %in% fanga18_zero))

which(colSums(ecm_hala12@otu_table) == 0)
which(colSums(ecm_hala18@otu_table) == 0) 
ecm_hala18_new = subset_samples(ecm_hala18, sample_id != "s72")

which(colSums(ecm_hala5p@otu_table) == 0)
ecm_hala5p_new <- subset_samples(ecm_hala5p, !(sample_id %in% c("s53", "s12")))

fruit12 <- fruit[, grep("2012", colnames(fruit))]
fruit18 <- fruit[, grep("2018", colnames(fruit))]
# changing 0 to NAs for chi test
fruit12[rowSums(fruit12) == 0, ] <- NA
fruit18 <- fruit18[!rowSums(fruit18) == 0, ]

# theorical values
## 
fruit12_th <- ifelse(fruit12 >= 0, 33.33333, 0)
fruit18_th <- ifelse(fruit18 >= 0, 33.33333, 0)
```

# *5 - VAR TRANSFO FOR ANALYSIS*

```{r transfo variables}
sam <- lapply(physeq@sam_data, as.factor) %>% as.data.frame()
sample_data(physeq)$year_factor <- as.factor(sample_data(physeq)$year)
```

```{r transposing otu table in physq object for hill_pq}
phy_otu_t <- t(physeq@otu_table)
phy_sam <- physeq@sam_data
phy_tax <- physeq@tax_table

ecm_phy_otu_t <- t(ecm_phy@otu_table) %>% otu_table()
ecm_phy_t <- phyloseq(ecm_phy_otu_t, ecm_phy@sam_data, ecm_phy@tax_table) %>% clean_pq()
```

```{r grouping by factor of insterest}
# grouping per level of treatment for raref curves
var1 <- as.character(get_variable(ecm_phy, "year"))
var2 <- as.character(get_variable(ecm_phy, "zone"))
sample_data(ecm_phy)$yr_treat <- mapply(paste0, var1, var2, collapse = "_")
phy_merged <- merge_group(ecm_phy, "yr_treat")

ecm_12 <- subset_samples(ecm_phy, ecm_phy@sam_data$year == "2012") %>% clean_pq(remove_empty_samples = FALSE,remove_empty_taxa = T, clean_samples_names = FALSE)
ecm_18 <- subset_samples(ecm_phy, ecm_phy@sam_data$year == "2018") %>% clean_pq(remove_empty_samples = FALSE,remove_empty_taxa = T, clean_samples_names = FALSE)

yr_12 <- merge_group(ecm_12, group = "zone",)
yr_18 <- merge_group(ecm_18, group = "zone",)
```

```{r diversity indexes}
# creating diversity variables
div <- vegan::diversity(x = t(ecm_phy@otu_table),
                             index = "shannon")
rich <- vegan::specnumber(x= t(ecm_phy@otu_table))
even <- div/ log(rich) # Pielou's evenness # tend to 1 = more evenly distributed 
tax_f$ID_TYPE[is.na(tax_f$ID_TYPE)] <- "AUTO_ID" 
even[even == 0] <- even["NaN"]

## relative abundance
# Agglomerate taxa at phylum level
phy_relat <- tax_glom(physeq, taxrank = "primary_lifestyle")

# Calculate relative abundance
phy_relat1 <- transform_sample_counts(phy_relat, function(x) x / sum(x))

# Subset object to only ECM
phy_relat2 <- subset_taxa(phy_relat1, primary_lifestyle == "ectomycorrhizal")
ecm_abun <- colSums(phy_relat2@otu_table) # total sum of ECM relative abundance / sample

## merging 
df_div <- data.frame(ecm_phy@sam_data, div, rich, even, ecm_abun) 
colnames(df_div)
df_div[, c(3:11)] <- lapply(df_div[, c(3:11)], as.factor) # tranforming columns into factors

# subsets
df_div_fanga <- subset(df_div, site == "Fangamon")
df_div_fanga12 <- subset(df_div_fanga, year == "2012")
df_div_fanga18 <- subset(df_div_fanga, year == "2018")
df_div_hala <- subset(df_div, site == "Halasen")
df_div_hala12 <- subset(df_div_hala, year == "2012")
df_div_hala18 <- subset(df_div_hala, year == "2018")
df_div_hala5p <- subset(df_div, site == "Halasen_p_5")
df_div12 <- subset(df_div, year == "2012")
df_div18 <- subset(df_div, year == "2018")

stde <- tapply(df_div$div, df_div$zone, stde) # if needed for tables
median <- tapply(df_div$div,df_div$zone, median)
min <- tapply(df_div$div,df_div$zone, min)
max <- tapply(df_div$div,df_div$zone, max)
```

```{r genus count}
ecm_phy1 <- ecm_phy
ecm_phy1 <- as_binary_otu_table(ecm_phy1)
osam <- data.frame(ecm_phy1@sam_data, t(ecm_phy1@otu_table))
colnames(osam)[c(12:156)] <- ecm_phy1@tax_table[, 6]

# count 
colnames(osam)
rownames(osam)

# count each occurence by group and year
osam_count <- osam[, c(4, 5, 9 ,14:length(colnames(osam)))] %>% 
  group_by(site, zone, year) %>% 
  summarise_all(sum)

# order df 
osam_count <- osam_count[
  with(osam_count, order(site, year, zone)),
] %>% as.data.frame()
rownames(osam_count) <- str_c(osam_count$site, osam_count$zone, osam_count$year, sep = "_")

# adding % occurence
## calc N
table(ecm_phy@sam_data$zone, ecm_phy@sam_data$year, ecm_phy@sam_data$site)
rownames(osam_count)

# create manually list (because too complicated otherwise)
N <- c(9, 9, 18, 9, 9, 9, 13, 8, 7, 15, 8, 7, 5, 11, 11)
osam <- data.frame(N, osam_count)

# Define the list of column names to calculate proportions for
colnames(osam)
otu_prop <- colnames(osam[, c(5: length(colnames(osam)))]) # Add all relevant column names here

# Apply the prop_fun() function to all columns and get a list of proportion vectors
prop_fun <- function(x) {
  x <- x / osam$N * 100
} # where:
## x = cell value; osam_count_df$N = the nb of samples for the corresponding row; * 100 = to get the proportion of genus occurrence by sample type. 
prop_list <- lapply(osam[, otu_prop], prop_fun)

# Convert the list of proportion vectors to a data frame
occurence_df <- as.data.frame(prop_list)
rownames(occurence_df) <- rownames(osam)
colnames(occurence_df) <- otu_prop
occurence_df <- occurence_df %>% round_df(1) # rounding

# subsetting per site 
occurence_fanga <- subset(occurence_df, grepl("Fangamon", rownames(occurence_df)))  
occurence_hala <- subset(occurence_df, !grepl("Halasen_p_5|Fangamon", rownames(occurence_df)))  
occurence_hala5p <- subset(occurence_df, grepl("Halasen_p_5", rownames(occurence_df)))  

# remove site from rowname, and only keep the treatment
rownames(occurence_fanga) <- gsub("Fangamon_", "", rownames(occurence_fanga))
rownames(occurence_hala) <- gsub("Halasen_", "", rownames(occurence_hala))
rownames(occurence_hala5p) <- gsub("Halasen_p_5_", "", rownames(occurence_hala5p))

# reordering rows
order_fanga <- factor(c("B", "C","A", "E", "F", "D")) # using letters ordination to order rows
occurence_fanga <- cbind(order_fanga, occurence_fanga)
occurence_fanga <- occurence_fanga[order(occurence_fanga$order), ]
occurence_fanga <- occurence_fanga[, c(2:ncol(occurence_fanga))] # removing ordering col

order_hala <- factor(c("B", "C","A", "E", "F", "D")) # using letters ordination to order rows
occurence_hala <- cbind(order_hala, occurence_hala)
occurence_hala <- occurence_hala[order(occurence_hala$order), ]
occurence_hala <- occurence_hala[, c(2:ncol(occurence_hala))] # removing ordering col

order_hala5p <- factor(c("B", "C","A")) # using letters ordination to order rows
occurence_hala5p <- cbind(order_hala5p, occurence_hala5p)
occurence_hala5p <- occurence_hala5p[order(occurence_hala5p$order), ]
occurence_hala5p <- occurence_hala5p[, c(2:ncol(occurence_hala5p))] # removing ordering col

# remove OTU that are from other sites
occurence_fanga <- subset(occurence_fanga, select = -which(colSums(occurence_fanga) == 0))
occurence_hala <- subset(occurence_hala, select = -which(colSums(occurence_hala) == 0))
occurence_hala5p <- subset(occurence_hala5p, select = -which(colSums(occurence_hala5p) == 0))

# order columns by OTU tot prop for each site
occurence_fanga1 <- rbind(occurence_fanga, totprop = colSums(occurence_fanga))
toccurence_fanga <- t(occurence_fanga1) %>% as.data.frame 
occurence_fanga2 <- toccurence_fanga[order(-toccurence_fanga$totprop), , drop = FALSE] 
occurence_fanga3 <- t(occurence_fanga2) %>% as.data.frame()
occurence_fanga <- occurence_fanga3[c(1:nrow(occurence_fanga-1)), ]

occurence_hala1 <- rbind(occurence_hala, totprop = colSums(occurence_hala))
toccurence_hala <- t(occurence_hala1) %>% as.data.frame 
occurence_hala2 <- toccurence_hala[order(-toccurence_hala$totprop), , drop = FALSE] 
occurence_hala3 <- t(occurence_hala2) %>% as.data.frame()
occurence_hala <- occurence_hala3[c(1:nrow(occurence_hala-1)), ]

occurence_hala5p1 <- rbind(occurence_hala5p, totprop = colSums(occurence_hala5p))
toccurence_hala5p <- t(occurence_hala5p1) %>% as.data.frame 
occurence_hala5p2 <- toccurence_hala5p[order(-toccurence_hala5p$totprop), , drop = FALSE] 
occurence_hala5p3 <- t(occurence_hala5p2) %>% as.data.frame()
occurence_hala5p <- occurence_hala5p3[c(1:nrow(occurence_hala5p-1)), ]

# deciding threshold
## 20 first 
ncol(occurence_fanga)
occurence_fanga2$totprop
ncol(occurence_hala)
occurence_hala2$totprop # most of least abundant OTU are composed of singletons (from to 13.3 and from 14.4, 14.3 being also a singleton)
## keeping the first 30 could be a good compromise (even if arbitrary)
ncol(occurence_hala5p)
occurence_hala5p2$totprop

## keeping 1st 30 most abundant OTU
occurence_fanga <- occurence_fanga[, c(0:40)]
occurence_hala <- occurence_hala[, c(0:40)]
occurence_hala5p <- occurence_hala5p[, c(0:40)]

pal_ht_fanga <- colorRampPalette(c("white", "cadetblue3", "darkmagenta"))(max(occurence_fanga))
pal_ht_hala <- colorRampPalette(c("white", "cadetblue3", "darkmagenta"))(max(occurence_hala))
pal_ht_hala5p <- colorRampPalette(c("white", "cadetblue3", "darkmagenta"))(max(occurence_hala5p))
pal_ht <- colorRampPalette(c("white", "cadetblue3", "darkmagenta"))(max(occurence_df))
```

# *6 - PALETTES - TESTS*

```{r dummy graphs}
list_mm <- lsf.str("package:MiscMetabar") 
factor(list_mm)
## to see all functions from a package

# circles physeq
circle_allsite_genus <- circle_pq(ecm_phy, "site", taxa = "Phylum", add_nb_seq = F,
          grid_col = "darkgrey",
          row_col = c("darkdarkred","#1F78B4","darkgreen"))

sample_names(ecm_phy)
```

```{r test piechart}
fortax <- data.frame(table(as.factor(ecm_phy@tax_table[,"Order"])))

fortax <- fortax %>% 
  group_by(Var1) %>% # Variable to be transformed
  count() %>% 
  ungroup() %>% 
  mutate(perc = `Freq` / sum(`Freq`)) %>% 
  arrange(perc) %>%
  mutate(labels = scales::percent(perc))

ggplot(fortax, aes(x ="", y = Freq, fill = Var1)) +
  geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0)+
  geom_text(aes(label = labels),
            position = position_stack(vjust = 0.5))
  
```

# *7 - MM RESULTS*

## DATA RUN

```{r MM manuscript}
# otu transformations
otu_zero_na <- otu_fungi
otu_zero_na[otu_zero_na==0] <- NA
mean_rd_sample <- colSums(otu_fungi) %>% mean() %>% print()
mean_rd_otu <- rowSums(otu_fungi) %>% mean() %>% print()

## total nb of reads
sum(otu_all) # total sum (all sites)
sum(otu_fungi_all) # only fungi (all sites)
sum(otu_fungi) # + hala
sum(otu_001) #  + <1% deleted # a lot because : most of OTUs have just a few reads. then, <1% of relative abundance per sample = deletion
sum(otu_rarefied) # + rarefaction
rd_del <- (sum(otu_fungi) - sum(otu_001)) %>% print()
raref_loss <- (sum(otu_001) - sum(otu_rarefied)) %>% print()

## loss of clusters 
nrow(otu_all) # total clusters (all sites)
nrow(otu_samples_out) # fungi + clusters in control (all sites)
nrow(otu_fungi_all) # fungi - clusters in control (all sites)
nrow(otu_fungi)
nrow(otu_001) #  + <1% deleted 
nrow(otu_rarefied) # + rarefaction
nrow(phy@otu_table)# + hala
ndel <- (nrow(otu_fungi) - nrow(otu_001)) %>% print()

plant_rd <-  (sum(otu_all) - sum(otu_fungi_all)) %>% print()
plant_otu <- (nrow(otu_samples_out) - nrow(otu_fungi_all) - length(clusters_in_controls)) %>% print()
clusters_in_controls  # clusters found in controls samples = suppressed
length(clusters_in_controls) %>% print()

## nb of taxa 
### OTUs 
summary(as.factor(taxeco1$ID_TYPE)) 
manual_id <- sum(as.factor(taxeco1$ID_TYPE) == "MANUAL_ID")
uncheck_otu <- sum(as.factor(taxeco1$ID_TYPE) == "AUTO_ID")
na_otu <- sum(is.na(taxeco1$OTU_ID))
auto_id <- uncheck_otu - na_otu

## IN HALA :

pourc_rd_hala <- (sum(otu_fungi)/sum(otu_fungi_all)*100) %>% print()

nb_rd_del <- (sum(otu_fungi) - sum(otu_001)) %>% print()
pourc_rd_del <- (sum(otu_001)/sum(otu_fungi)*100) %>% print()

nb_otu_del <- (nrow(otu_fungi) - nrow(otu_001)) %>% print() 

nrow(phy@tax_table)
summary(as.factor(phy@tax_table[, "primary_lifestyle"]))
# N OTU
n_otu_ecm <- nrow(ecm_phy@tax_table) %>% print() # ECM
n_otu_sapr<- nrow(sapr_phy@tax_table) %>% print() # SAPR
n_otu_other <- nrow(other_phy@tax_table) %>% print() # OTHER 

# N read
nb_rd_ecm <- sum(ecm_phy@otu_table) %>% print() # ECM
nb_rd_sapr<- sum(sapr_phy@otu_table) %>% print() # SAPR
nb_rd_other <- sum(other_phy@otu_table) %>% print() # OTHER  

# % read (rarefied)
pourc_ecm <- (nb_rd_ecm/(nb_rd_ecm + nb_rd_sapr + nb_rd_other)*100) %>% print() 
pourc_sapr <- (nb_rd_sapr/(nb_rd_ecm + nb_rd_sapr + nb_rd_other)*100) %>% print() 
pourc_other <- (nb_rd_other/(nb_rd_ecm + nb_rd_sapr + nb_rd_other)*100) %>% print() 
```

```{r results mm}
colnames(run_data) 

run_data[,1]# 2 runs
run_data[,25] # 98,5% similarity clustering

# run 1
run_data[1,2] # total nb of pyro reads
run_data[1,3] # sequences passing quality control (QC)
run_data[1,2]-run_data[1,3] # sequences deleted

# run 2
run_data[2,2] # total nb of pyro reads
run_data[2,3] # sequences passing quality control (QC)
run_data[2,2]-run_data[2,3] # sequences deleted

# scata taxa attribution 

## no IDed sequences
is.na(raw_clust[,4]) %>% summary() # 1145 clusters ID / 4050 
pourc_id <- 1145/4050*100; pourc_id # autoID (especially the most abundant clusters)
# % ID to be considered in regard of the % of total reads. 
pourc_id_read <- 

## NF sequences
grep("PLANT|NF", raw_clust[,4], value = TRUE) %>% length() # 82 non-fungi clusters IDed 

# total VS conserve
## nb clusters
nrow(raw_clust) # all
nrow(ecm_phy@tax_table) # ecm among 1000st more abundant Fungi clusters 
pourc_ecm <- nrow(ecm_phy@tax_table)/1000*100; pourc_ecm # 20% ECM guild among Fungi

## nb reads
run_data[3,2]
raw_plantfungi_read <- sum(raw_clust$Cluster_Size) 
raw_total_read <- sum(raw_f_clust$Cluster_Size) # total read of fungi clusters 
ecm_total_read <-sum(ecm_raw$Cluster_Size) # only the 1st 1000 clusters are analyzed through Funguild to define their trophic state.

pourc_nb_read_fungi <- (raw_total_read/raw_plantfungi_read)*100; pourc_nb_read_fungi
pourc_nb_read_ecm <- ecm_total_read/raw_total_read*100 %>% print() # pourc nb read conserved/ nb read 
summary(ecm_raw$Cluster_Size) # min, mean, max of ecm reads
prop_read_ID <- sum(taxeco_id$Cluster_Size)/sum(raw_f_clust$Cluster_Size)*100
prop_read_ID # 93.62% of reads from sequencing are from the 1000 first clusters manually verified (and kept for further analysis)
prop_read_001 <-   # proportion of seq. reads kept after supressing the <1% reads / sample. 
```

## OVERVIEW_DATA

```{r general}
summary(run_data)
summary(raw_clust)
summary(ecm_raw)
summary(sapr_raw)
levels(droplevels(ecm_raw$Genus))
levels(droplevels(sapr_raw$Genus))

summary(ecm_raw$Cluster_Size)

hist_cluster_size <- ggplot(ecm_raw, 
       aes(x=Cluster_Size))+
geom_histogram(bins=20,breaks=seq(12,2985 ,by=100),color="black", fill="darkseagreen")+
  scale_y_continuous(name="number of clusters")+
  scale_x_continuous(name="number of reads", breaks=seq(12,2985 ,by=100))+
  theme(axis.text.x=element_text(color="coral4", size=8, angle=60))+
  ggtitle("Distribution of the cluster size for ECM OTUs") 

hist_cluster_size

read_distribution <- plot_read_distribution(ecm_phy, groups = "site", 
                            plot.type = "density")
read_distribution
print(read_distribution + theme_biome_utils())
```

```{r SAC nb read}
sac_f_raw <- ggplot(raw_f_clust, aes(x=c(1:3968),y = cumsum(Cluster_Size)))+   geom_line(color="darkred")+
  geom_point(size=0.7,alpha=0.5)+
  coord_cartesian(xlim=c(0,3968),ylim=c(0,244367))+
  scale_y_continuous(name="cumulative cluster size", breaks=seq(0,244367,by=25000))+
  scale_x_continuous(name="number of cluster", breaks=seq(0,3968 ,by=200))+
  geom_vline(xintercept=1000,lwd=0.6,colour="darkred",)+ 
  annotate("text", x = 1050, y = 25000, label="1000 first clusters conserved", size = 4, angle=0, color = "darkred", hjust = 0)+
  annotate("text", x= 1050, y= 15000, label="93.65% of total reads", size = 4, angle=0, color = "darkred", hjust = 0)+
  theme(axis.text.x=element_text(color="coral4", size=8, angle=60))+
  ggtitle("Cumulative cluster size curve from all fungi clusters") 

sac_f_raw

read_1000 <- sum(raw_f_clust[c(1:1000),2]); read_1000
read_3968 <- sum(raw_f_clust[c(1001:3968),2]); read_3968
read_total <- sum(raw_f_clust[,2])
pourc_read_kept <- read_1000/read_total*100; pourc_read_kept
```

```{r sample desing}

nrow(otu) # total number of all clusters (ie fungi ecm or no-ecm, plants)
colnames(otu)

# samples missing post scata (ie not present in the OTU matrix from clustering = no match between ITS4 - ITS7 tags)

samples_missing_postscata <- outersect(x = sam_no_match$samples_match, y = sam_no_match$samples_all) # antijoin but for columns : extract samples not in common (missing from initial samples design)
samples_missing_postscata <- dput(as.character(samples_missing_postscata)) # convert into the format c(x; y; ...; z)
sam[sam$sample_id %in% samples_missing_postscata,] # showing samples with site and plots information missing 
## 5 controls; 3 Halasen; 2 Halasen_p5, 1 Q1; 6 PCR_neg
## total missing = 17

sample_count_design <- length(rownames(sam)) # 178 samples total design + controls)

# design samples per site 
length(rownames(sam_hala))
length(rownames(sam_hala5))
length(rownames(sam_fanga))

# sample count
length(rownames(sam_controls_na)) # 19 (12 NAs + 7 controls)
sum(is.na(sam_controls_na$site)) 
real_sample_count <- (length(rownames(sam))-length(rownames(sam_controls_na))) %>%
  print() # 159 sites samples


colnames(otu)
s_id_controls <- sam_controls_na$sample_id[sam_controls_na$site == "control"] %>%
  na.exclude()  # what are the samples_id of controls samples ? 
s_id_controls <- dput(as.character(s_id_controls))

col_control_kept <- which(colnames(otu) %in% s_id_controls, arr.ind = F) 
colnames(otu)[col_control_kept]

controls_kept <- c("sITS-standard2", "sITS-standard1", "sPCR_neg_1", "sPCR_neg_2", "sQ2", "s49", "s97") # 2 ITS; 2 PCR; 1Q; 2 controls 

161-7+5 
# 161 = total samples kept after scata
# -7  = controls kept after scata
# +5  = site samples missing 
## final 159 samples -> ALL GOOD 
```

```{r missing samples}
# considering physeq objects (ecm_phy,ecm_phy, hala5_ecm, fanga_ecm have been cleaned (sample supress if 0) by clean_pq() )
colnames(hala_sam)
summary(hala_sam[,c(2,5:8)])
table(hala_sam$plot,hala_sam$treatment)

# for ecm-sapr
summary(sam$sample_id)
summary(physeq@sam_data$sample_id)
miss_sam_clust_ecmsapr <- anti_join(x = sam, y = physeq@sam_data) # return all rows from x without a match in y.
miss_sam_clust_ecmsapr$sample_id # samples deleted after bioinformatic processes (scata, first 1000s clusters, ecm-sapr guilds)

# for ecm only
summary(sam$sample_id)
summary(ecm_phy@sam_data$sample_id)
miss_sam_clust_ecm <- anti_join(x = sam, y = ecm_phy@sam_data)
miss_sam_clust_ecm$sample_id
# 8 PCR neg => good control,
# what about ITS standard ?

clipr::write_clip(miss_sam_clust_ecm$sample_id) # to copy dataframe or vector
# controls and na 
miss_sam_clust_ctrl <- anti_join(x = sam_controls_na, y = ecm_phy@sam_data)
miss_sam_clust_ctrl$sample_id

length(miss_sam_clust_ctrl$sample_id)
length(row.names(sam_controls_na)) # every control and NA are suppressed from the dataset (because 0 oSCurence of clusters)

# hala
miss_sam_clust_hala <- anti_join(x = sam_hala[,1:7], y = ecm_phy@sam_data[,1:7]) # exclude zone column because we changed factor values for this column
miss_sam_clust_hala$sample_id

# hala5
miss_sam_clust_hala5 <- anti_join(x = sam_hala5, y = hala5_ecm@sam_data)
miss_sam_clust_hala5$sample_id

# fanga
miss_sam_clust_fanga <- anti_join(x = sam_fanga, y = fanga_ecm@sam_data)
miss_sam_clust_fanga$sample_id

# check if count is good (total missing samples VS sum of missing samples per site)
length(miss_sam_clust_ecm$sample_id)/(length(miss_sam_clust_hala$sample_id)+length(miss_sam_clust_hala5$sample_id)+length(miss_sam_clust_fanga$sample_id)+length(miss_sam_clust_ctrl$sample_id)) 
# =1 -> all good 

sam[sam$sample_id %in% samples_missing_postscata,] # showing samples with site and plots information missing 
## 5 controls; 3 Halasen; 2 Halasen_p5, 1 Q1; 6 PCR_neg
## total missing = 17
length(miss_sam_clust_ecm$sample_id) # total 
length(miss_sam_clust_ctrl$sample_id)

length(miss_sam_clust_hala$sample_id) # 3 missing post scata; 3 missing post taxa-attrib
length(miss_sam_clust_hala5$sample_id)# 2 missing post scata; 2 missing post taxa-attrib
length(miss_sam_clust_fanga$sample_id)# 0 missing post scata; 20 missing post taxa-attrib

length(rownames(sam_hala))
length(rownames(sam_hala5))
length(rownames(sam_fanga))
```

```{r potential contamination ?}
which(samples_kept_pq>0, arr.ind = T) # location where values of controls have reads
clusters_reads_inctrl <- rownames(which(samples_kept_pq>0, arr.ind = T)) %>% as.factor()
levels(clusters_reads_inctrl) # clusters where we found reads in controls samples 
## -> should we erase them for dataset ? 
## surely, imo 
```

```{r rarefaction and hellinger transformation comparison}
###### distribution
all_distrib_raref <- plot_read_distribution(
  physeq, groups = "site", 
  plot.type = "density")+
  xlab("Reads per site")+
  ggtitle("Distribution of the number of reads per site")
all_distrib_raref

all_distrib_notraref <- plot_read_distribution(
  physeq_notrarefied, groups = "site", 
  plot.type = "density")+
  xlab("Reads per site")+
  ggtitle("Distribution of the number of reads per site")
all_distrib_notraref

all_distrib_relative <- plot_read_distribution(
  physeq_relative, groups = "site", 
  plot.type = "density")+
  xlab("Reads per site")+
  ggtitle("Distribution of the number of reads per site")
all_distrib_relative

all_distrib_hellinger <- plot_read_distribution(
  physeq_hellinger, groups = "site", 
  plot.type = "density")+
  xlab("Reads per site")+
  ggtitle("Distribution of the number of reads per site")
all_distrib_hellinger

####### raref 
all_curve_raref <- phyloseq.extended::ggrare(physeq = physeq, 
                                             step = 10, 
                                             color = "year", 
                                             label="sample_id",
                                             plot=F, 
                                             se=T)+ 
  facet_wrap(~site)+
  ggtitle("Rarefaction curves of samples per site")+
  xlab("Number of reads")+
  ylab("OTU richness")
all_curve_raref

all_curve_raref_group <- phyloseq.extended::ggrare(physeq, step = 10, color = "site",plot=F, se=T, label = "sample_id")+ 
  labs(x = "Group size",
       y =  "OTU richness",
       title = "rarefaction curves by samples",
       color = "Treatment")+
  guides(fill = F) # Suppressing the filling (standard error) legend 
all_curve_raref_group

all_curve_notraref_group <- phyloseq.extended::ggrare(physeq_notrarefied, step = 10, color = "site",plot=F, se=T, label = "sample_id")+ 
  labs(x = "Group size",
       y =  "OTU richness",
       title = "rarefaction curves by samples",
       color = "Treatment")+
  guides(fill = F) # Suppressing the filling (standard error) legend 
all_curve_notraref_group

all_curve_notraref_group <- phyloseq.extended::ggrare(physeq_relative, step = 10, color = "site",plot=F, se=T, label = "sample_id")+ 
  labs(x = "Group size",
       y =  "OTU richness",
       title = "rarefaction curves by samples",
       color = "Treatment")+
  guides(fill = F) # Suppressing the filling (standard error) legend 
all_curve_notraref_group

######### multivariate
## hellinger
all_mdsbray_hellinger <- ordinate(physeq_hellinger, method = "MDS", dist="bray")
all_mdsbray_hellinger <- plot_ordination(physeq_hellinger, all_mdsbray_hellinger, type="samples", color="site",label="sample_id")+
  geom_point(size=2)+
  facet_wrap(~year)+
  # color and elipse
  scale_color_manual(values = c("Halasen_p_5" = "blue3",
                              "Halasen" = "green4",
                              "Fangamon" = "darkorange"))+
  stat_ellipse(type="t", level=0.70)+
  # legend
  guides(color=guide_legend(title="position to SC"))+
  ggtitle(label = "MDS of samples to Halasen_edge by year", subtitle="presence-absence (brayard distance)")
all_mdsbray_hellinger

## justabundance
all_mdsbray_relativeab <- ordinate(physeq_relative, method = "MDS", dist="bray")
all_mdsbray_relativeab <- plot_ordination(physeq_relative, all_mdsbray_relativeab, type="samples", color="site",label="sample_id")+
  geom_point(size=2)+
  facet_wrap(~year)+
  # color and elipse
  scale_color_manual(values = c("Halasen_p_5" = "blue3",
                              "Halasen" = "green4",
                              "Fangamon" = "darkorange"))+
  stat_ellipse(type="t", level=0.70)+
  # legend
  guides(color=guide_legend(title="position to SC"))+
  ggtitle(label = "MDS of samples to Halasen_edge by year", subtitle="presence-absence (brayard distance)")
all_mdsbray_relativeab

## not rarefied
all_mdsbray_notrarefied <- ordinate(physeq_notrarefied, method = "MDS", dist="bray")
all_mdsbray_notrarefied <- plot_ordination(physeq_notrarefied, all_mdsbray_notrarefied, type="samples", color="site",label="sample_id")+
  geom_point(size=2)+
  facet_wrap(~year)+
  # color and elipse
  scale_color_manual(values = c("Halasen_p_5" = "blue3",
                              "Halasen" = "green4",
                              "Fangamon" = "darkorange"))+
  stat_ellipse(type="t", level=0.70)+
  # legend
  guides(color=guide_legend(title="position to SC"))+
  ggtitle(label = "MDS of samples from Halasen_edge by year", subtitle="presence-absence (brayard distance)")
all_mdsbray_notrarefied

## rarefied
all_mdsbray_rarefied <- ordinate(physeq, method = "MDS", dist="bray")
all_mdsbray_rarefied <- plot_ordination(physeq, all_mdsbray_rarefied, type="samples", color="site",label="sample_id")+
  geom_point(size=2)+
  facet_wrap(~year)+
  # color and elipse
  scale_color_manual(values = c("Halasen_p_5" = "blue3",
                              "Halasen" = "green4",
                              "Fangamon" = "darkorange"))+
  stat_ellipse(type="t", level=0.70)+
  # legend
  guides(color=guide_legend(title="position from SC"))+
  ggtitle(label = "MDS of samples from Halasen_edge by year", subtitle="presence-absence (brayard distance)")
all_mdsbray_rarefied
```

```{r check losing clusters - samples after clean_pq}
# nb of clusters in subsets
nrow(otu_all)             # 4050 
nrow(otu1)                # 3411
nrow(physeq@otu_table)    # 588
nrow(ecm_phy@otu_table)   # 149
nrow(sapr_phy@otu_table)  # 172 
nrow(fanga_ecm@otu_table) # 37
nrow(ecm_phy@otu_table)  # 100
nrow(hala5_ecm@otu_table) # 49

# representation of the total nb of reads
sum(otu_all) # total all reads
sum(otu_samples_out) # total reads with controls out 
sum(otu_fungi) # total reads only fungi

sum(otu_fungi)/sum(otu_all)*100 # 75% fungi in all reads 

sum(otu1)-sum(otu_001) # losing clusters with <1% relative abundance = 81668 reads
sum(otu_001) # without <1% clusters 

sum(physeq_notrarefied@otu_table)
sum(physeq_notrarefied@otu_table)-sum(physeq@otu_table) # loosing 115 997 reads with rarefaction
sum(physeq@otu_table) # 40 502 reads after rarefaction in all sites 

sum(ecm_phy@otu_table) # 8 379 ecm reads
sum(sapr_phy@otu_table)# 18 615 saprotroph reads
sum(other_phy@otu_table)# 13508 reads from "other" (other guilds and not attributed, could be saprobes) 

# per guild - all sites
pourc_ecm <- sum(ecm_phy@otu_table)/sum(physeq@otu_table)*100; pourc_ecm # 21% ecm 
pourc_sapr <- sum(sapr_phy@otu_table)/sum(physeq@otu_table)*100; pourc_sapr # 46 % sapro
pourc_other <- sum(other_phy@otu_table)/sum(physeq@otu_table)*100; pourc_other # 33 % unknown (potential sapro?)
pourc_other + pourc_sapr + pourc_ecm

# per site (rarefied)
sum(ecm_phy@otu_table)
sum(hala5_ecm@otu_table)
sum(fanga_ecm@otu_table)

pourc_ecm_phy <- sum(ecm_phy@otu_table)/sum(hala@otu_table)*100; pourc_ecm_phy # 29.2% ecm hala
pourc_hala5_ecm <- sum(hala5_ecm@otu_table)/sum(hala5@otu_table)*100; pourc_hala5_ecm # 23.1 % ecm hala5  
pourc_fanga_ecm <- sum(fanga_ecm@otu_table)/sum(fanga@otu_table)*100; pourc_fanga_ecm # 12.6% fanga
```

```{r sequencing depth changing between pool ?}
richness_allc_raref <- specnumber(t(otu_rarefied)) # richness calc (same as nonrarefied ofc) 
summary(richness_allc_raref) # richness of samples from ALL Fungi clusters
# hist(richness_allc_raref) # PLOT 

reads_sum_per_sample <- data.frame(rownames(otu_t), rowSums(otu_t)) # after deleting <1% relative abundance
summary(reads_sum_per_sample) # sum of reads per sample for ALL Fungi clusters 
shapiro.test(reads_sum_per_sample$rowSums.otu_t.) # normal distribution
#hist(reads_sum_per_sample$rowSums.otu_t.) # PLOT # rarefying based on the min sample depth (here 419)
low_seqdepth_samples <- which(reads_sum_per_sample < 300, arr.ind = T) %>% data.frame() %>% print() # samples with sequencing depth under a threshold
list_low_seqdepth_samples <- dput(as.character(rownames(low_seqdepth_samples))) # convert into the format c(x; y; ...; z)
# TOASK -> should we delete these samples ? 
# View(sam[list_low_seqdepth_samples,])
## check pool 1 2 sequencing depth difference

physeq_notrarefied_p1 <- subset_samples(physeq_001kept, physeq_001kept@sam_data$pool=="1") %>% clean_pq(remove_empty_samples = T,remove_empty_taxa = T, clean_samples_names = T)

physeq_notrarefied_p2 <- subset_samples(physeq_001kept, physeq_001kept@sam_data$pool=="2") %>% clean_pq(remove_empty_samples = T,remove_empty_taxa = T, clean_samples_names = T)

r_sps <- rowSums(t(physeq_001kept@otu_table)) # r_sps = sum per sample
summary(r_sps)
#hist(r_sps) #PLOT

rp1_sps <- rowSums(t(physeq_notrarefied_p1@otu_table))
summary(rp1_sps)
pool1 <- 1
rp1_sps_df <- data.frame("pool" = 1, "sum_reads" = rp1_sps)

rp2_sps <- rowSums(t(physeq_notrarefied_p2@otu_table)) 
summary(rp2_sps)
rp2_sps_df <- data.frame("pool" = 2, "sum_reads" = rp2_sps)

r_sps_df <- bind_rows(rp1_sps_df, rp2_sps_df)
r_sps_df <- cbind(physeq_001kept@sam_data$year, physeq_001kept@sam_data$plot, r_sps_df)
colnames(r_sps_df) <- c("year", "plot", "pool", "sum_reads")
r_sps_df$pool <- as.factor(r_sps_df$pool)

# graphical representation
mu_pool <- ddply(r_sps_df, "pool", summarise, grp.mean=mean(sum_reads)) # calculed mean of each group 
ggplot(r_sps_df, aes(x = sum_reads, color = pool, fill = pool)) +
  geom_histogram(aes(y = ..density..), alpha = 0.5, position = "identity")+
  geom_density(alpha=0.3)+
  geom_vline(data = mu_pool, aes(xintercept=grp.mean, color= pool),
           linetype="dashed")

boxplot(r_sps ~ physeq_001kept@sam_data$pool)
boxplot(r_sps ~ physeq_001kept@sam_data$plot)


rd_sps_inf1_gg <- reshape2::melt(rd_sps_inf1[,2:3])
ggplot(rd_sps_inf1_gg, aes(x = value, color = variable, fill = variable))+
  geom_histogram(aes(y = ..count..), alpha = 0.4, position = "identity")+
  geom_density(alpha = 0.3)+
  scale_y_continuous(name = "nb of samples")+
  scale_x_continuous(name = "read sum per sample", breaks = seq(0,3000 ,by = 200))+
  ggtitle("Number of samples per read sum")+
  geom_vline(xintercept = c(mean(rd_sps_inf1$sumreads_otu1), mean(rd_sps_inf1$sumreads_otu_001)),lwd = 0.6,colour = c("red","blue"))

# statistical test
lm_seq_depth <- lmer(sum_reads ~ pool + (pool | plot) + (pool | year), r_sps_df)
qqnorm(residuals(lm_seq_depth))
qqline(residuals(lm_seq_depth))
plot(lm_seq_depth)
summary(lm_seq_depth)
anova(lm_seq_depth)  # if test conditions arent fulfilled 
# TOASK seems to be significant difference in seq depth for the pools
```

# *8 - RESULTS PAPER*

```{r GENERAL}
table(ecm_phy@sam_data$zone, ecm_phy@sam_data$year, ecm_phy@sam_data$site) # desing post sub-sampling, 1% otu suppresion, etc.

read_distrib <- plot_read_distribution(
  ecm_phy, groups = "site", 
  plot.type = "density")+
  xlab("Reads per site")+
  ggtitle("Distribution of the number of reads per site")
read_distrib
```

```{r COMPOSITION}
asc <- which(sum_res$data == "Ascomycota") %>% length()
bas <- which(sum_res$data == "Basidiomycota") %>% length()
pourc_asc <- (asc / (asc + bas))*100; pourc_asc
pourc_bas <- (bas / (asc + bas))*100; pourc_bas
```

```{r ALPHA_DIV}
bxplt_rich <- df_div %>%
  ggplot(aes(x = factor(zone, levels = treat_order), y = rich, fill = factor(zone)))+
  geom_boxplot(width = 0.4, color = "black") +
  coord_cartesian(ylim=c(0,15))+
  scale_y_continuous(name = "OTU richness", breaks = seq(0, 15, by = 2))+
  theme(axis.text.x=element_text(color="coral4", size = 11, angle = 90))+
  facet_grid(~ site*year) +
  theme(legend.position="none", plot.title = element_text(size = 14)) +
  labs(title = "ECM OTU richness by cutting zone")+
  xlab("Zone")+
  guides(fill = guide_legend(title = str_wrap("ECM OTU richness by cutting zone", width = 20)))+
  scale_fill_manual(values = treat_color)
bxplt_rich

bxplt_relat <- df_div %>%
  ggplot(aes(x = factor(zone, levels = treat_order), y = ecm_abun, fill = factor(zone)))+
  geom_boxplot(width = 0.4, color = "black") +
  coord_cartesian(ylim=c(0,1))+
  scale_y_continuous(name = "OTU richness", breaks = seq(0, 1, by = 0.2))+
  theme(axis.text.x=element_text(color="coral4", size = 11, angle = 90))+
  facet_grid(~ site*year) +
  theme(legend.position="none", plot.title = element_text(size = 14)) +
  labs(title = "ECM OTU relative abundance by cutting zone")+
  xlab("Zone")+
  guides(fill = guide_legend(title = str_wrap("ECM OTU relative abundance by cutting zone", width = 20)))+
  scale_fill_manual(values = treat_color)
bxplt_relat
```

```{r BETA DIV}
# Fangamon
bray_fanga12_ord <- ordinate(physeq = ecm_fanga12_new, method = "MDS", dist="bray")
bray_fanga12 <- plot_ordination(physeq = ecm_fanga12_new, ordination = bray_fanga12_ord, type = "samples", color = "zone")+
  geom_point(size = 2)+
  geom_text(label = ecm_fanga12_new@sam_data$sample_id, size = 2.7, nudge_y = +0.05)+
  scale_color_manual(values = coltreatment_saturated, breaks = treat_order)+
  stat_ellipse(type="t", level=0.70)+
  guides(color=guide_legend(title="Zone"))+
    ggtitle(label = "MDS of samples grouped by zone (F1_12)",subtitle = "by number of sequence reads (bray-curtis index)")
bray_fanga12

bray_fanga18_ord <- ordinate(physeq = ecm_fanga18_new, method = "MDS", dist="bray")
bray_fanga18 <- plot_ordination(physeq = ecm_fanga18_new, ordination = bray_fanga18_ord, type = "samples", color = "zone")+
  geom_point(size = 2)+
  geom_text(label = ecm_fanga18_new@sam_data$sample_id, size = 2.7, nudge_y = +0.05)+
  scale_color_manual(values = coltreatment_saturated, breaks = treat_order)+
  stat_ellipse(type="t", level=0.70)+
  guides(color=guide_legend(title="Zone"))+
    ggtitle(label = "MDS of samples grouped by zone (F1_18)",subtitle = "by number of sequence reads (bray-curtis index)")
bray_fanga18

# Halasen
bray_hala12_ord <- ordinate(physeq = ecm_hala12, method = "MDS", dist="bray")
bray_hala12 <- plot_ordination(physeq = ecm_hala12, ordination = bray_hala12_ord, type = "samples", color = "zone")+
  geom_point(size = 2)+
  geom_text(label = ecm_hala12@sam_data$sample_id, size = 2.7, nudge_y = +0.05)+
  scale_color_manual(values = coltreatment_saturated, breaks = treat_order)+
  stat_ellipse(type="t", level=0.70)+
  guides(color=guide_legend(title = "Zone"))+
  ggtitle(label = "MDS of samples grouped by zone (H1_12)",subtitle = "by number of sequence reads (bray-curtis index)")+
  theme(text = element_text(family = "Arial"))
bray_hala12

bray_hala18_ord <- ordinate(physeq = ecm_hala18_new, method = "MDS", dist="bray")
bray_hala18 <- plot_ordination(physeq = ecm_hala18_new, ordination = bray_hala18_ord, type = "samples", color = "zone")+
  geom_point(size = 2)+
  geom_text(label = ecm_hala18_new@sam_data$sample_id, size = 2.7, nudge_y = +0.05)+
  scale_color_manual(values = coltreatment_saturated, breaks = treat_order)+
  stat_ellipse(type="t", level=0.70)+
  guides(color=guide_legend(title="Zone"))+
    ggtitle(label = "MDS of samples grouped by zone (H1_18)",subtitle = "by number of sequence reads (bray-curtis index)")
bray_hala18

# Halasen 5p
bray_hala5p_ord <- ordinate(physeq = ecm_hala5p_new, method = "MDS", dist="bray")
bray_hala5p <- plot_ordination(physeq = ecm_hala5p_new, ordination = bray_hala5p_ord, type = "samples", color = "zone")+
  geom_point(size = 2)+
  geom_text(label = ecm_hala5p_new@sam_data$sample_id, size = 2.7, nudge_y = +0.05)+
  scale_color_manual(values = coltreatment_saturated, breaks = treat_order)+
  stat_ellipse(type="t", level=0.70)+
  guides(color=guide_legend(title="Zone"))+
    ggtitle(label = "MDS of samples grouped by zone (H2)",subtitle = "by number of sequence reads (bray-curtis index)")
bray_hala5p
```

```{r MODEL RICH}
# Fangamon
# verif assumption for glmm
summary(df_div_fanga$rich)
summary(df_div_fanga$ecm_abun)
par(mfrow = c(1, 2))
hist(df_div_fanga$rich, breaks = seq(0, 15, by = 1))
hist(df_div_fanga$ecm_abun, breaks = seq(0, 1, by = 0.1))
hist(log(ecm_abun))
par(mfrow = c(1, 1))

glmm_rich_fanga <- lme4::glmer(rich ~ zone * year +
                           (1 | plot),
                         family = poisson(link = "log"),
                         data = df_div_fanga) 

summary(glmm_rich_fanga)
car::Anova(glmm_rich_fanga, type = "III")
pair_glmm_rich_fanga <- emmeans(glmm_rich_fanga, ~ zone, type = "response")
pairs(pair_glmm_rich_fanga)

# 2012 & 2018
glmm_rich_fanga12 <- lme4::glmer(rich ~ zone +
                           (1 | plot),
                         family = poisson(link = "log"),
                         data = df_div_fanga12) 
glmm_rich_fanga18 <- lme4::glmer(rich ~ zone +
                           (1 | plot),
                         family = poisson(link = "log"),
                         data = df_div_fanga18)

summary(glmm_rich_fanga12)
car::Anova(glmm_rich_fanga12, type = "III")
tab_model(glmm_rich_fanga12, show.icc = F, show.df = F, show.ngroups = F, show.re.var = T, show.ngroups = F, show.re.var = T)
pair_glmm_rich_fanga12 <- emmeans(glmm_rich_fanga12, ~ zone, type = "response")
pairs(pair_glmm_rich_fanga12)

summary(glmm_rich_fanga18)
car::Anova(glmm_rich_fanga18, type = "III")
tab_model(glmm_rich_fanga18, show.icc = F, show.df = F, show.ngroups = F, show.re.var = T)
pair_glmm_rich_fanga18 <- emmeans(glmm_rich_fanga18, ~ zone, type = "response")
pairs(pair_glmm_rich_fanga18)

# Halasen SC
# verif assumption for glmm
summary(df_div_hala$rich)
summary(df_div_hala$ecm_abun)
par(mfrow = c(1, 2))
hist(df_div_hala$rich, breaks = seq(0, 15, by = 1))
hist(df_div_hala$ecm_abun, breaks = seq(0, 1, by = 0.1))
hist(log(ecm_abun))
par(mfrow = c(1, 1))

glmm_rich_hala <- lme4::glmer(rich ~ zone * year +
                           (1 | plot),
                         family = poisson(link = "log"),
                         data = df_div_hala) 

summary(glmm_rich_hala)
car::Anova(glmm_rich_hala, type = "III")
pair_glmm_rich_hala <- emmeans(glmm_rich_hala, ~ zone, type = "response")
pairs(pair_glmm_rich_hala)

# 2012 & 2018
glmm_rich_hala12 <- lme4::glmer(rich ~ zone +
                           (1 | plot),
                         family = poisson(link = "log"),
                         data = df_div_hala12) 
glmm_rich_hala18 <- lme4::glmer(rich ~ zone +
                           (1 | plot),
                         family = poisson(link = "log"),
                         data = df_div_hala18)

summary(glmm_rich_hala12)
car::Anova(glmm_rich_hala12, type = "III")
tab_model(glmm_rich_hala12, show.icc = F, show.df = F, show.ngroups = F, show.re.var = T)
pair_glmm_rich_hala12 <- emmeans(glmm_rich_hala12, ~ zone, type = "response")
pairs(pair_glmm_rich_hala12)

summary(glmm_rich_hala18)
car::Anova(glmm_rich_hala18, type = "III")
tab_model(glmm_rich_hala18, show.icc = F, show.df = F, show.ngroups = F, show.re.var = T)
pair_glmm_rich_hala18 <- emmeans(glmm_rich_hala18, ~ zone, type = "response")
pairs(pair_glmm_rich_hala18)

# Halasen 5p
# verif assumption for glmm
summary(df_div_hala5p$rich)
summary(df_div_hala5p$ecm_abun)
par(mfrow = c(1, 2))
hist(df_div_hala5p$rich, breaks = seq(0, 15, by = 1))
hist(df_div_hala5p$ecm_abun, breaks = seq(0, 1, by = 0.1))
hist(log(ecm_abun))
par(mfrow = c(1, 1))

glmm_rich_hala5p <- lme4::glmer(rich ~ zone +
                           (1 | plot),
                         family = poisson(link = "log"),
                         data = df_div_hala5p) 

summary(glmm_rich_hala5p)
car::Anova(glmm_rich_hala5p, type = "III")
tab_model(glmm_rich_hala5p, show.icc = F, show.df = F, show.ngroups = F, show.re.var = T)
pair_glmm_rich_hala5p <- emmeans(glmm_rich_hala5p, ~ zone, type = "response")
pairs(pair_glmm_rich_hala5p)
```

```{r MODEL ABUN}
# Fangamon
# verif assumption for glmm
summary(df_div_fanga$ecm_abun)
par(mfrow = c(1, 2))
hist(df_div_fanga$ecm_abun, breaks = seq(0, 1, by = 0.1))
hist(log(ecm_abun))
par(mfrow = c(1, 1))

glmm_abun_fanga <- lme4::glmer(ecm_abun ~ zone * year +
                           (1 | plot),
                         family = poisson(link = "log"),
                         data = df_div_fanga) 

summary(glmm_abun_fanga)
car::Anova(glmm_abun_fanga, type = "III")
pair_glmm_abun_fanga <- emmeans(glmm_abun_fanga, ~ zone, type = "response")
pairs(pair_glmm_abun_fanga)

# 2012 & 2018
glmm_abun_hala18 <- lme4::glmer(ecm_abun ~ zone +
                           (1 | plot),
                         family = poisson(link = "log"),
                         data = df_div_hala18) 
glmm_abun_fanga18 <- lme4::glmer(ecm_abun ~ zone +
                           (1 | plot),
                         family = poisson(link = "log"),
                         data = df_div_fanga18)

summary(glmm_abun_hala18)
car::Anova(glmm_abun_hala18, type = "III")
pair_glmm_abun_hala18 <- emmeans(glmm_abun_hala18, ~ zone, type = "response")
pairs(pair_glmm_abun_hala18)

summary(glmm_abun_fanga18)
car::Anova(glmm_abun_fanga18, type = "III")
pair_glmm_abun_fanga18 <- emmeans(glmm_abun_fanga18, ~ zone, type = "response")
pairs(pair_glmm_abun_fanga18)

# Halasen SC
# verif assumption for glmm
summary(df_div_hala$ecm_abun)
summary(df_div_hala$ecm_abun)
par(mfrow = c(1, 2))
hist(df_div_hala$ecm_abun, breaks = seq(0, 15, by = 1))
hist(df_div_hala$ecm_abun, breaks = seq(0, 1, by = 0.1))
hist(log(ecm_abun))
par(mfrow = c(1, 1))

glmm_abun_hala <- lme4::glmer(ecm_abun ~ zone * year +
                           (1 | plot),
                         family = poisson(link = "log"),
                         data = df_div_hala) 

summary(glmm_abun_hala)
car::Anova(glmm_abun_hala, type = "III")
pair_glmm_abun_hala <- emmeans(glmm_abun_hala, ~ zone, type = "response")
pairs(pair_glmm_abun_hala)

# 2012 & 2018
glmm_abun_hala12 <- lme4::glmer(ecm_abun ~ zone +
                           (1 | plot),
                         family = poisson(link = "log"),
                         data = df_div_hala12) 
glmm_abun_hala18 <- lme4::glmer(ecm_abun ~ zone +
                           (1 | plot),
                         family = poisson(link = "log"),
                         data = df_div_hala18)

summary(glmm_abun_hala12)
car::Anova(glmm_abun_hala12, type = "III")
pair_glmm_abun_hala12 <- emmeans(glmm_abun_hala12, ~ zone, type = "response")
pairs(pair_glmm_abun_hala12)

summary(glmm_abun_hala18)
car::Anova(glmm_abun_hala18, type = "III")
pair_glmm_abun_hala18 <- emmeans(glmm_abun_hala18, ~ zone, type = "response")
pairs(pair_glmm_abun_hala18)

# Halasen 5p
# verif assumption for glmm
summary(df_div_hala5p$ecm_abun)
summary(df_div_hala5p$ecm_abun)
par(mfrow = c(1, 2))
hist(df_div_hala5p$ecm_abun, breaks = seq(0, 15, by = 1))
hist(df_div_hala5p$ecm_abun, breaks = seq(0, 1, by = 0.1))
hist(log(ecm_abun))
par(mfrow = c(1, 1))

glmm_abun_hala5p <- lme4::glmer(ecm_abun ~ zone +
                           (1 | plot),
                         family = poisson(link = "log"),
                         data = df_div_hala5p) 

summary(glmm_abun_hala5p)
car::Anova(glmm_abun_hala5p, type = "III")
pair_glmm_abun_hala5p <- emmeans(glmm_abun_hala5p, ~ zone, type = "response")
pairs(pair_glmm_abun_hala5p)
```

```{r MODEL FANGAMON}
# verif assumption for glmm
summary(df_div_fanga$rich)
summary(ecm_abun_df_fanga$ecm_abun)
par(mfrow = c(1, 2))
hist(df_div_fanga$rich, breaks = seq(0, 15, by = 1))
hist(ecm_abun_df_fanga$ecm_abun, breaks = seq(0, 1, by = 0.1))
hist(log(ecm_abun))
par(mfrow = c(1, 1))

# GLMM richness 

# general & red
glmm_rich_fanga <- lme4::glmer(rich ~ zone * year +
                           (1 | plot),
                         family = poisson(link = "log"),
                         data = df_div_fanga) 

glmm_richred_fanga <- lme4::glmer(rich ~ (1 | plot),
                         family = poisson(link = "log"),
                         data = df_div_fanga) 

summary(glmm_rich_fanga)
car::Anova(glmm_rich_fanga, type = "III")
anova(glmm_rich_fanga, glmm_richred_fanga) # likely-hood ration test (LRT test) # where dff is the number of parameters that vary between the 2 models
summary(glmm_rich)
pair_glmm_rich_fanga <- emmeans(glmm_rich_fanga, ~ zone, type = "response")
pairs(pair_glmm_rich_fanga)
tab_model(glmm_rich_fanga, show.df = F, show.icc = F)

# 2012 & 2018
glmm_rich_fanga12 <- lme4::glmer(rich ~ zone +
                           (1 | plot),
                         family = poisson(link = "log"),
                         data = df_div_fanga12) 
glmm_rich_fanga18 <- lme4::glmer(rich ~ zone +
                           (1 | plot),
                         family = poisson(link = "log"),
                         data = df_div_fanga18)

summary(glmm_rich_fanga12)
car::Anova(glmm_rich_fanga12, type = "III")
pair_glmm_rich_fanga12 <- emmeans(glmm_rich_fanga12, ~ zone, type = "response")
pairs(pair_glmm_rich_fanga12)
tab_model(glmm_rich_fanga12, show.df = F, show.icc = F)

summary(glmm_rich_fanga18)
car::Anova(glmm_rich_fanga18, type = "III")
pair_glmm_rich_fanga18 <- emmeans(glmm_rich_fanga18, ~ zone, type = "response")
plot(pair_glmm_rich_fanga18)
pairs(pair_glmm_rich_fanga18)

# NS general model, 
# when considered years independantly: S for year 2018, between FOR - CUT but NS between TH_B - TH_NB 
##  consider to merge both to increase statistical power
```

```{r HEATMAP}
braydist = phyloseq::distance(ecm_phy, "bray")
dist_df = reshape2::melt(as.matrix(braydist))

# grouped heatmaps
## by treatment 

#### breaks (continuous or discreet ?)
fanga_breaks <- seq(min(occurence_fanga), max(occurence_fanga), length.out = 10)
fanga_breaks

hala_breaks <- seq(min(occurence_hala), max(occurence_hala), length.out = 10)
hala_breaks

hala5p_breaks <- seq(min(occurence_hala5p), max(occurence_hala5p), length.out = 10)
hala5p_breaks

# full heatmap
htmap_fanga <- pheatmap::pheatmap(mat = t(occurence_fanga), 
                   cellwidth = 30, cellheight = 7,
                   cluster_row = F, cluster_col = F,
                   fontsize_row = 7, fontsize_col = 10,
                   color = pal_ht,
                   annotation_col = df_color,
                   annotation_colors = c(anot_colors_treat, anot_colors_year),
                   main = "% OTU occurence by zone (F1)",
                   display_numbers = F, 
                   number_format = "%.0f", 
                   fontsize_number = 5,
                   na_col = "white")

htmap_hala <- pheatmap::pheatmap(mat = t(occurence_hala), 
                   cellwidth = 30, cellheight = 7,
                   cluster_row = F, cluster_col = F,
                   fontsize_row = 7, fontsize_col = 10,
                   color = pal_ht,
                   annotation_col = df_color,
                   annotation_colors = c(anot_colors_year, anot_colors_treat),
                   main = "% OTU occurence by zone (H1)")

htmap_hala5p <- pheatmap::pheatmap(mat = t(occurence_hala5p), 
                   cellwidth = 30, cellheight = 7,
                   cluster_row = F, cluster_col = F,
                   fontsize_row = 7, fontsize_col = 10,
                   color = pal_ht,
                   annotation_col = df_color,
                   annotation_colors = c(anot_colors_year, anot_colors_treat),
                   main = "% OTU occurence by zone (H2)")

### 
ht_count0 <- htmp_all$data %>% group_by(OTU_ID, yr_treat) %>%
  dplyr::summarise(sumcount = sum(Abundance, na.rm = TRUE)) %>% as.data.frame()

ht_count <- ht_count0 %>%
  # convert state to factor and reverse order of levels
  mutate(Genus = factor(Genus, levels = rev(sort(unique(Genus))))) %>%
  # create a new variable from count
  mutate(countfactor = cut(sumcount, breaks = c(-1, 1, 5, 10, 50, 150, 450, max(sumcount, na.rm = T)),
                          labels=c("0", "0-1", "5-10", "10-50", "50-150", "150-450", ">450"))) %>%
  # change level order
  mutate(countfactor=factor(as.character(countfactor), levels=rev(levels(countfactor))))

### trying to creat multi xlabel
yr <- ifelse(grepl("2012", ht_count$yr_treat), "2012", "2018")
treat <- ifelse(grepl("CCFM", ht_count$yr_treat), "CCFM",
                         ifelse(grepl("SC", ht_count$yr_treat), "SC", "FOR")) # nested ifelse
ht_count <- cbind(ht_count, yr, treat)
### dont know.

## plotting
htmap_all <- ggplot(ht_count,  aes(x = factor(yr_treat, levels = order_level_mg), y = Genus, fill = countfactor)) + 
  geom_tile(colour = "white", size = 0.4)+
  scale_fill_manual(values = rev(c("grey90", RColorBrewer::brewer.pal(n = 5, name = "YlOrRd"))), na.value = "grey90")+
  theme_grey(base_size = 10)+
  guides(fill =guide_legend(title = str_wrap("Number of sequence reads (rarefied)", width = 20)))+
  labs(x = "Distance to cut and year", y = "Genus", title = "Occurence of OTU Genus by distance to cut and by year")+
  scale_x_discrete(labels = order_level_simpl)

htmap_all
```

# *9 - H1) HALASEN STUDY: strip-cut effect*

## 3.0 design and sampling

```{r overview }
# sampling design
colnames(sam)
summary(sam[,c(2,5:8)])

# post biomol design
table_postbiomol <- table(sam_hala$zone, sam_hala$plot) # post biomol
table_postbiomol <- table_postbiomol[, plot_list] %>% print()
rowSums(table_postbiomol) ## N = 3 missing in initial sample_data matrix (64 samples in design) missing after the biomolecular analysis
### plots: 2 (FOR, n=1); 9 (FOR, n=2)
n_postbiomol <- length(sam_hala$sample_id) %>% print()
# balanced now after resampling of forest 

# post bioinf design
table_postbioinf <- table(ecm_phy@sam_data$zone, ecm_phy@sam_data$plot) # post bioinf
table_postbioinf <- table_postbioinf[, plot_list] %>% print()
rowSums(table_postbioinf) ## after clean_pq 
## 2 (n=1); 9 (n=2); 3 (n=3); 8 (n=1) -> N= 4 
n_postbionf <- length(ecm_phy@sam_data$sample_id) %>% print()

# not present in otu_table:
sam_hala[which(rownames(sam_hala) %in% rownames(otu_fungi_t0) == FALSE), ]
sam_hala$sample_id[which(rownames(sam_hala) %in% rownames(otu_fungi_t0) == FALSE)] %>% print()
## samples missing from otu table (during clustering scata)
### is this a pb in tag matching ? 

Ntreat12
Ntreat18

# 0 ECM sample: 
which(colSums(ecm_phy@otu_table) == 0) 
ecm_phy@sam_data[which(colSums(ecm_phy@otu_table) == 0)]
## sample with 0 ECM read 

## --> total missing : 4 + 3 = 7 (64 - 7 = 57)

# design table
summary(as.factor(sam$treatment)) 
summary(as.factor(ecm_phy@sam_data$zone))

tableall <- table(ecm_phy@sam_data$zone, ecm_phy@sam_data$plot) %>% print() %>% as.data.frame.matrix() 

tableall <- tableall[, plot_list] %>% print() # reordering column as similar as design

table12 <- table(ecm_12@sam_data$zone, ecm_12@sam_data$plot) %>% as.data.frame.matrix()
table12 <- table12[, plot_list] %>% print() # not working bc not all forest plot are present anymore after resampling

table18 <- table(ecm_18@sam_data$zone, ecm_18@sam_data$plot) %>% as.data.frame.matrix()
table18 <- table18[, plot_list] %>% print()

sum(table12)+sum(table18) # verif ok
Nall <- rowSums(tableall) %>% print()
Nyear12 <- sum(table12) %>% print()
Nyear18 <- sum(table18) %>% print()
Ntreat12 <- rowSums(table12) %>% print()
Ntreat18 <- rowSums(table18) %>% print()

# nb genus
colnames(ecm_phy@tax_table)
as.factor(ecm_phy@tax_table[,5]) %>% droplevels() %>% levels() 
genus_nb_hala <- as.factor(ecm_phy@tax_table[,5]) %>% droplevels() %>% levels() %>% length()
## 27 genus
## attribution facolors, or whatever to have proper constrasted colors on graphicals representations

##### checking cluster nb etc
phytax_df <- data.frame(phy@tax_table) %>% lapply(as.factor) %>% data.frame()
rownames(phytax_df) <- rownames(phy@tax_table)
str(phytax_df)
summary_all <- summary(phytax_df) %>% print()
summary_all_df <- data.frame(unclass(summary(phytax_df)), check.names = FALSE, stringsAsFactors = FALSE)

str(ecmtax_df)
summary_ecm <- summary(ecmtax_df) %>% print()
summary_ecm_df <- data.frame(unclass(summary(ecmtax_df)), check.names = FALSE, stringsAsFactors = FALSE)

level_order_ecmtax <- c("Phylum", "Class", "Family", "Genus", "Fruitbody")

# should be better if arrange in proper order

colnames(ecm_phy@otu_table)
colSums(ecm_phy@otu_table) %>% summary()
# total changes from colsums just after rarefaction because not_ecm clusters are supressed
colSums(phy@otu_table) %>% summary 

hist(colSums(ecm_phy@otu_table))

## clusters--OTU in control
clusters_in_controls
```

```{r summary results}
sum_res <- ecmtax_df[, c("Phylum", "Class", "Genus", "Family", "Fruitbody")] %>%
  gather(variable, value) %>%
  ggplot()+
  geom_bar(aes(x = factor(variable, levels = level_order_ecmtax), y = ..count../sum(..count..)*5, fill = value), color= " black", position = "stack", show.legend = FALSE)+
  geom_text(stat = "count", aes(x = variable, label =  value,  y = ..count../sum(..count..)*5, group = value), position = "stack", vjust = 1, alpha = 1)+
  scale_y_continuous(labels = scales::percent)+
  ylab("") # 5 because of the nb of variables
sum_res
```

-   change between 2012 2018 (see below) -\> could be explained by a
    difference in the biomolecular methods ? we should not expect any
    difference in forest control treatment between years

## 3.1 stat test

```{r indexes model treatment * year }
summary(rich)
summary(div)
summary(even)

str(df_div)

bxplt_rich <- df_div %>%
  ggplot(aes(x = factor(zone, levels = order_level), y = rich, fill =  zone))+
  geom_boxplot(width = 0.4, color = "black") +
  coord_cartesian(ylim=c(0,15))+
  scale_y_continuous(name = "OTU richness", breaks = seq(0, 15, by = 2))+
  theme(axis.text.x=element_text(color="coral4", size = 11, angle = 0))+
  facet_grid(~ year) +
  theme(legend.position="none", plot.title = element_text(size = 14)) +
  labs(title = "ECM OTU richness by distance from cut")+
  xlab("Distance from cut")+
  guides(fill = guide_legend(title = str_wrap("Distance from cut", width = 20)))+
  scale_fill_manual(values = coltreatment)
bxplt_rich

table_hala <- table(df_div_hala$zone, df_div_hala$plot) %>% as.data.frame.matrix()
table_hala
# variation 
mean12 <- aggregate(df_div12$rich, by = list(df_div12$zone), FUN = mean) %>% print()
sd12 <- aggregate(df_div12$rich, by = list(df_div12$zone), FUN = sd) %>% print()

mean18 <- aggregate(df_div18$rich, by = list(df_div18$zone), FUN = mean) %>% print()
sd18 <-  aggregate(df_div18$rich, by = list(df_div18$zone), FUN = sd) %>% print()


summary(as.factor(df_div$zone))
```

```{r relative abundance}
genus_nb_sapr <- as.factor(sapr_phy@tax_table[,5]) %>% droplevels() %>% levels() %>% length() %>% print()
genus_nb_ecm <- as.factor(ecm_phy@tax_table[,5]) %>% droplevels() %>% levels() %>% length() %>% print()
## OTU Genus count (0;1) by treatment
phy_ggplot1 <- phyloseq::plot_bar(as_binary_otu_table(physeq), x = "zone") 

phy_ggplot1 <- ggplot(phy_ggplot1$data, aes(fill= fct_reorder(as.factor(primary_lifestyle), Abundance), x = factor(zone, levels = order_level), y = Abundance))+
  geom_bar(stat = "identity", position = "stack") +
  facet_grid(~year, scale="free")+
  labs(title="OTU Genus count (0;1) by position from SC and year")+
  xlab("position from SC")+
  guides(fill=guide_legend(title="Genus"))+
  theme(axis.text.x=element_text(color="coral4", size=8, angle=20))
phy_ggplot1

bxplt_relat <- df_div %>% ggplot(aes(x = factor(zone, levels = order_level), y = ecm_abun, fill = zone))+
  geom_boxplot(width = 0.4, color = "black") +
  coord_cartesian(ylim=c(0,1))+
  scale_y_continuous(name = "ECM proportion", breaks = seq(0, 1, by = 0.1))+
  theme(axis.text.x=element_text(color="coral4", size = 11, angle = 0))+
  facet_grid(~ year) +
  theme(legend.position = "none", plot.title = element_text(size = 14)) +
  scale_fill_manual(values = coltreatment)+
  labs(title = "ECM proportion by distance from cut")+
  xlab("Distance from cut")
bxplt_relat

# variation 
sd12_prop <- aggregate(ecm_abun_df12$ecm_abun, by = list(ecm_abun_df12$zone), FUN = sd) %>% print()
sd18_prop <-  aggregate(ecm_abun_df18$ecm_abun, by = list(ecm_abun_df18$zone), FUN = sd) %>% print()
mean(ecm_abun_df18$ecm_abun[ecm_abun_df18$zone == "CCFM"])

mean12_prop <- aggregate(ecm_abun_df12$ecm_abun, by = list(df_div12$zone), FUN = mean) %>% print()
mean18_prop <- aggregate(ecm_abun_df18$ecm_abun, by = list(df_div18$zone), FUN = mean) %>% print()
```

```{r MODEL }
# https://stats.stackexchange.com/questions/31569/questions-about-how-random-effects-are-specified-in-lmer
## for constructing model with random effects

## for interaction : 
## ":" or "*" -> figure which symbol is appropriate

# spatial autocorrelation -> lags ? to consider into glmm. is that also possible to perform glmm if autocorrelation anyways ?
## using ape package , moran's index 
# if so, correct 

# not possible to put in random effect a variable that is NA for levels of other variable (eg stream with CCFM from treatment)
# glmer non hypothesis of normality, so just plot(model) .> fitted values by 

# verif assumption for glmm
summary(rich)
summary(ecm_abun)
par(mfrow = c(1, 2))
hist(rich, breaks = seq(0, 15, by = 1))
hist(ecm_abun, breaks = seq(0, 1, by = 0.1))
hist(log(ecm_abun))
par(mfrow = c(1, 1))

# GLMM richness 
glmm_rich <- lme4::glmer(rich ~ zone * year +
                           (1 | plot),
                         family = poisson(link = "log"),
                         data = df_div) 
glmm_richred <- lme4::glmer(rich ~ (1 | plot),
                         family = poisson(link = "log"),
                         data = df_div) 

car::Anova(glmm_rich, type = "III")
anova(glmm_rich, glmm_richred) # likely-hood ration test (LRT test) # where dff is the number of parameters that vary between the 2 models
summary(glmm_rich)
pair_glmm_rich <- emmeans(glmm_rich, ~ zone * year, type = "response")
pairs(pair_glmm_rich)
tab_model(glmm_rich, show.df = TRUE)

# GLMM richness 2012
glmm_rich12 <- lme4::glmer(rich ~ zone +
                           (1 | plot),
                         family = poisson(link = "log"),
                         data = df_div12) 
glmm_rich12red <- lme4::glmer(rich ~ (1 | plot),
                         family = poisson(link = "log"),
                         data = df_div12) 

car::Anova(glmm_rich12, type = "III")
anova(glmm_rich12, glmm_rich12red)
summary(glmm_rich12)
drop1(glmm_rich12, test = "Chisq")
lrtest(glmm_rich12)
pair_glmm_rich12 <- emmeans(glmm_rich12, ~ zone, type = "response")
pairs(pair_glmm_rich12)

# GLMM richness 2018
glmm_rich18 <- lme4::glmer(rich ~ zone +
                           (1 | plot),
                         family = poisson(link = "log"),
                         data = df_div18) 
glmm_rich18b <- lme4::glmer(rich ~
                           (1 | plot),
                         family = poisson(link = "log"),
                         data = df_div18) 

car::Anova(glmm_rich18, type = "III")
drop1(glmm_rich18)
anova(glmm_rich18, glmm_rich18red)
summary(glmm_rich18)
lrtest(glmm_prop18)
pair_glmm_rich18 <- emmeans(glmm_rich18, ~ zone, type = "response")
pairs(pair_glmm_rich18)

# GLMM prop
glmm_prop <- glmmTMB::glmmTMB(ecm_abun ~ zone * year +
                        (1 | plot),
                        family = beta_family(link = "logit"),
                        data = ecm_abun_df) 
car::Anova(glmm_prop, type = "III")
summary(glmm_prop)
pair_glmm_prop <- emmeans(glmm_prop, ~ zone * year, type = "response")
pairs(pair_glmm_prop)

# GLMM prop 2012
glmm_prop12 <- glmmTMB::glmmTMB(ecm_abun ~ zone +
                        (1 | plot),
                        family = beta_family(link = "logit"),
                        data = ecm_abun_df12)

car::Anova(glmm_prop12, type = "III")
summary(glmm_prop12)
pair_glmm_prop12 <- emmeans(glmm_prop12, ~ zone, type = "response")
pairs(pair_glmm_prop12)

# GLMM prop 2018
glmm_prop18 <- glmmTMB::glmmTMB(ecm_abun ~ zone +
                        (1 | plot),
                        family = beta_family(link = "logit"),
                        data = ecm_abun_df18) 

car::Anova(glmm_prop18, type = "III")
summary(glmm_prop18)
pair_glmm_prop18 <- emmeans(glmm_prop18, ~ zone, type = "response")
pairs(pair_glmm_prop18)


pchisq(summary(glm_rich)$deviance, summary(glm_rich)$df.residual) # test for testing fitting of family 

par(mfrow = c(2, 2))
plot(glm_rich) # graphical assumption tests
par(mfrow = c(1, 1))

# plot fixed effects correlation matrix
sjPlot::plot_model(glmm_rich)
# plot qq-plot of random effects
sjp.glmer(fit, type = "re.qq")


# analysis of variance (lm) or deviance (glm and glmm)
tableall
Nyear12; Ntreat12; Nyear18; Ntreat18
summary(glmm_rich) # 1 
# summary interpretation of fixed effects: 
## intercept = CCFM + 2012 + CCFM:2012, necessarily significant because model no null (0)
## SC and FOR are S=/= of CCFM independently of year     NB: (S=/= means significantly different)
## year 2018 is =/= from 2012 independently of treatment
## SC:2018 (ie in interaction) is =/= from CCFM:2012 -> SC and CCFM difference do vary through years with an observed "inversment" where richness of the edges decrease through time when richness of middle is increasing.
## FOR:2018 is =/= from CCFM:2012 -> FOR and CCFM difference do not vary through years (a global decrease for both of treatments is observed)

# summary interpretation of correlations:

anova(glmm, test = "LR") # order influence significance intercept (CCFM select as intercept, alphabetically or 1st row of df.)
summary(glmm_richtest) # 2
anova(glmm_richtest)
summary(glm) # 3
anova(glm, test = "LR") # ou methode des moindres carres, etc. # function # anova (analysis of deviance) is working only for glm, not glmm
summary(lm)
anova(lm)

anova(glm, test = "LR")
## copied from dataset
all.ai=manyglm(allotus.ai ~ as.numeric(Reads_mean) + Pair + as.numeric(AI), data=factors[-11,], family="negative.binomial")
plot(all.ai)
anova.all.ai=anova(all.ai, test="LR", resamp="perm.resid", p.uni="adjusted")

# TABLE 4b #
anova.all.ai

which(anova.all.ai$uni.p[2,]<=0.05)
which(anova.all.ai$uni.p[3,]<=0.05)
which(anova.all.ai$uni.p[4,]<=0.05)
##

predict(glmm_want)
plot(glmm_want) # distribution res 
VarCorr(glmm_want)

#calculate McFadden's R-squared for model
with(summary(glm), 1 - deviance/null.deviance)

# pairwise comparisons
tuk <- emmeans(glm, "zone")
contrast(tuk, "pairwise", adjust = "Tukey")

emmeans(glm, pairwise ~ zone : year) # possible to do pairwise on interaction
emmeans(lm, pairwise ~ zone)
TukeyHSD(aov(lm))

##
# assumption test for model
hist(rich) # not normal : poisson distribution lambda = 2 

boxplot(rich ~ zone, data = df_div)

par(mfrow=c(1, 3))
hist(rich_) # graphical distrib
hist(residuals(glmm_want))
hist(residuals(lm_rich))
shapiro.test(glmm_want) # shapiro tests
shapiro.test(residuals(glmm_want))
shapiro.test(residuals(lm_rich))

par(mfrow=c(2, 2))
plot(glmm_rich_pois)
qqnorm(residuals(glmm_want))
qqline(residuals(glmm_want))

par(mfrow=c(1,1))

shapiro.test(lm_rich$res) # OK
dwtest(lm_rich)  # OK
bptest(lm_rich)  # OK
anova(lm_rich)
summary(lm_rich)
TukeyHSD(aov(lm_rich), "specification_value")

summary(glmm_rich_pois)


boxplot(rich_ ~ ecm_phy@sam_data$specification_value)

# model all addition ## TO RESOLVE AND CHOOSE
str(df_div)

lm_div <- lm(rich_ ~ orient + stream, data = df_div) 

par(mfrow=c(2,2))
plot(lm_div)
par(mfrow=c(1,1))
hist(lm_div$res)
shapiro.test(lm_div$res) # OK
dwtest(lm_div)  # OK
bptest(lm_div)  # OK
Anova(lm_div, type="II")
anova(lm_div)
TukeyHSD(aov(lm_div), "specification_value")

boxplot(rich_hala ~ ecm_phy@sam_data$specification_value)
```

## 3.2.1 alpha TY

```{r rarecurve}
# maybe consider the nb of samples (to limit biais in unbalanced design
min(rowSums(yr_12@otu_table))
hala_rrcurve_groups1 <- phyloseq.extended::ggrare(yr_12, step = 10, color = "zone",plot=F, se=T)+ 
  facet_wrap(~ year)+
  coord_cartesian(ylim = c(1, max(vegan::specnumber(yr_12@otu_table))))+
  scale_x_continuous(breaks = seq(0, max(rowSums(yr_12@otu_table)), by = 200))+
  scale_y_continuous(breaks = seq(0, max(vegan::specnumber(yr_12@otu_table)), by = 5))+
  geom_vline(xintercept = min(rowSums(yr_18@otu_table)), color = "coral4")+
  labs(x = "Group size",
       y =  "OTU richness",
       title = "Rarefaction curves by distance from cut",
       color = "Treatment")+
  theme(legend.position = c(0.9, 0.6))+
  guides(fill = F, title = str_wrap("Distance from cut", width = 20))+
  annotate("text", x = 190, y = 4, label = str_wrap("n = 179", width = 20), size = 3.5, angle=0, color = "coral4", hjust = 0)+
  scale_color_manual(values = coltreatment_raref)+
  scale_fill_manual(values = coltreatment_raref)
hala_rrcurve_groups1
# Group size = (ie sum of samples reads per group)

min(rowSums(yr_18@otu_table))
hala_rrcurve_groups2 <- phyloseq.extended::ggrare(yr_18, step = 10, color = "zone",plot=F, se=T)+ 
  facet_wrap(~year)+
  geom_vline(xintercept = min(rowSums(yr_18@otu_table)), color = "coral4")+
  coord_cartesian(ylim = c(1, max(vegan::specnumber(yr_18@otu_table))))+
  scale_x_continuous(breaks = seq(0, max(rowSums(yr_18@otu_table)), by = 200))+
  scale_y_continuous(breaks = seq(0, max(vegan::specnumber(yr_18@otu_table)), by = 5))+
  labs(x = "Group size",
       y =  "OTU richness",
       title = "Rarefaction curves by distance from cut",
       color = "Treatment")+
  guides(fill = F)+ # Suppressing the filling (standard error) legend 
  theme(legend.position = c(0.85, 0.6))+
  annotate("text", x = 190, y = 2, label = str_wrap("n = 179", width = 20), size = 3.5, angle=0, color = "coral4", hjust = 0)+
  scale_color_manual(values = coltreatment_raref)+
  scale_fill_manual(values = coltreatment_raref)
hala_rrcurve_groups2

hala_rrcurve <- phyloseq.extended::ggrare(ecm_phy, step = 10, color = "zone", plot = F, se=T)+
  labs(x = "Group size",
       y =  "OTU richness",
       title = "rarefaction curves by treatment",
       color = "Treatment")
hala_rrcurve
```

```{r circle graph: genus distribution among treatments}
circle_pq(physeq = ecm_phy, 
                              fact = "zone", 
                              taxa = "Genus", 
                              min_prop_mod=0.001,
                              min_prop_tax=0.01,
                              row_col = c("SC" = "darkred",
                                          "CCFM" = "orange",
                                          "FOR" = "darkgreen"),
                              grid_col= 1,
                              add_nb_seq = T)
```

```{r alpha diversity indexes on otu}
plot_richness(ecm_phy, measures = c("Chao1", "ACE", "InvSimpson"), nrow=3, sortby = "Chao1", x = "zone")

hala_hill_treatment <- hill_pq(physeq = ecm_phy_t, 
                               variable = "zone",
                               color_fac = NA, 
                               letters = F)
  

hala_hill0 <- hala_hill_treatment[[1]]+ 
  labs(title="Hill 0 : Distance from cut", subtitle="Hagadalen_edge")+
  coord_cartesian(xlim=c(0,16))+
  scale_x_continuous(name="Hill_0", breaks = seq(0,16,by = 2))+
  scale_y_continuous(name="Distance from SC", breaks=c(1,2,3))+
  scale_color_manual(values = c("SC" = "darkred",
                                "CCFM" = "orange",
                                "FOR" = "darkgreen"))+
  theme(axis.text.x=element_text(color="coral4", size=8, angle=0))
hala_hill0

hala_hill1 <- hala_hill_treatment[[2]]+ 
  labs(title="Hill 1 : Distance from cut", subtitle="Hagadalen_edge")+
  coord_cartesian(xlim=c(0,12))+
  scale_x_continuous(name="Hill_1", breaks=seq(0,12,by=2))+
  scale_y_continuous(name="Distance from SC", breaks=c(1,2,3))+
  scale_color_manual(values = c("CCFM" = "orange",
                                "SC" = "darkred",
                                "FOR" = "darkgreen"))+
  theme(axis.text.x=element_text(color="coral4", size=8, angle=0))
hala_hill1

hala_hill2 <- hala_hill_treatment[[3]]+ 
  labs(title="Hill 2 : Distance from SC", subtitle="Hagadalen_edge")+
  coord_cartesian(xlim=c(0,10))+
  scale_x_continuous(name="Hill_2", breaks=seq(0,10,by=2))+
  scale_y_continuous(name="Distance from SC", breaks=c(1,2,3))+
  scale_color_manual(values = c("CCFM" = "orange",
                                "SC" = "darkred",
                                "FOR" = "darkgreen"))+
  theme(axis.text.x=element_text(color="coral4", size=8, angle=0))
hala_hill2

hala_hill_treatment[[4]]
hala_hill_treatment[[4]] # TukeyHSD test
```

```{r barplot genus distribution - occurence}
genus_nb_hala <- as.factor(ecm_phy@tax_table[,5]) %>% droplevels() %>% levels() %>% length() %>% print()
## OTU Genus count (0;1) by treatment
ecm_phy_ggplot1 <- phyloseq::plot_bar(as_binary_otu_table(ecm_phy), x = "zone") 
ecm_phy_ggplot1 <- ggplot(ecm_phy_ggplot1$data, aes(fill= fct_reorder(as.factor(Genus), Abundance), x = factor(zone, levels = order_level), y = Abundance))+
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = funky(genus_nb_hala))+
  facet_grid(~year, scale="free")+
  labs(title="OTU Genus presence-absence by distance from cut and by year")+
  xlab("Distance from cut")+
  ylab("OTU Genus count")+
  guides(fill=guide_legend(title="Genus"))+
  theme(axis.text.x = element_text(color="black", size=8, angle=20))+
  theme(legend.key.size = unit(0.4, 'cm'), #change legend key size
        legend.key.height = unit(0.4, 'cm'), #change legend key height
        legend.key.width = unit(0.4, 'cm'), #change legend key width
        legend.title = element_text(size = 11), #change legend title font size
        legend.text = element_text(size = 9)) #change legend text font size
ecm_phy_ggplot1

# OTU Genus abundance by treatment
ecm_phy_ggplot2 <- phyloseq::plot_bar(ecm_phy, x = "zone") 
ecm_phy_ggplot2 <- ggplot(ecm_phy_ggplot2$data, aes(fill= fct_reorder(as.factor(Genus), Abundance), x = factor(zone, levels = order_level), y = Abundance))+
  geom_bar(stat="identity", position = "stack")+
  scale_fill_manual(values = funky(genus_nb_hala))+
  facet_grid(~year, scale="free")+
  labs(title="OTU Genus abundance by distance from cut and by year")+
  xlab("Distance from cut")+
  ylab("Number of seq. reads")+
  guides(fill=guide_legend(title="Genus"))+
  theme(axis.text.x=element_text(color="black", size=8, angle=20))+
  theme(legend.key.size = unit(0.4, 'cm'), #change legend key size
        legend.key.height = unit(0.4, 'cm'), #change legend key height
        legend.key.width = unit(0.4, 'cm'), #change legend key width
        legend.title = element_text(size = 11), #change legend title font size
        legend.text = element_text(size = 9)) #change legend text font size
ecm_phy_ggplot2
```

```{r barplot genus distribution - proportion}
# tirage aleatoire pour nb d'echantillon (bootstrap ?) to compensate effect of Nforest 

genus_nb_hala <- as.factor(ecm_phy@tax_table[,5]) %>% droplevels() %>% levels() %>% length() %>% print()
## OTU Genus count (0;1) by treatment
ecm_phy_ggplot1_prop <- phyloseq::plot_bar(as_binary_otu_table(ecm_phy), x = "zone") 
ecm_phy_ggplot1_prop <- ggplot(ecm_phy_ggplot1_prop$data, aes(fill= fct_reorder(as.factor(Genus), Abundance), x = factor(zone, levels = order_level), y = Abundance))+
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = funky(genus_nb_ecm))+
  facet_grid(~year, scale="free")+
  labs(title="OTU Genus presence-absence by distance from cut and by year")+
  xlab("Distance from cut")+
  ylab("OTU Genus count")+
  guides(fill=guide_legend(title="Genus"))+
  theme(axis.text.x = element_text(color="black", size=8, angle=20))+
  theme(legend.key.size = unit(0.4, 'cm'), #change legend key size
        legend.key.height = unit(0.4, 'cm'), #change legend key height
        legend.key.width = unit(0.4, 'cm'), #change legend key width
        legend.title = element_text(size = 11), #change legend title font size
        legend.text = element_text(size = 9)) #change legend text font size
ecm_phy_ggplot1_prop
Ntreat12
Ntreat18

# OTU Genus abundance by treatment
ecm_phy_ggplot2_prop <- phyloseq::plot_bar(ecm_phy, x = "zone") 
ecm_phy_ggplot2_prop <- ggplot(ecm_phy_ggplot2_prop$data, aes(fill= fct_reorder(as.factor(Genus), Abundance), x = factor(zone, levels = order_level), y = Abundance))+
  geom_bar(stat="identity", position = "fill")+
  scale_fill_manual(values = funky(genus_nb_hala))+
  facet_grid(~year, scale="free")+
  labs(title="OTU Genus proportion by distance from cut and by year")+
  xlab("Distance from cut")+
  ylab("OTU Genus proportion")+
  guides(fill=guide_legend(title="Genus"))+
  theme(axis.text.x=element_text(color="black", size=8, angle=20))+
  theme(legend.key.size = unit(0.4, 'cm'), #change legend key size
        legend.key.height = unit(0.4, 'cm'), #change legend key height
        legend.key.width = unit(0.4, 'cm'), #change legend key width
        legend.title = element_text(size = 11), #change legend title font size
        legend.text = element_text(size = 9)) #change legend text font size
ecm_phy_ggplot2_prop
```

```{r barplot OTUs distribution}
otu_nb_hala <- as.factor(ecm_phy@tax_table[,6]) %>% droplevels() %>% levels() %>% length()
## OTU count (0;1) by treatment
ecm_phy_ggplot3 <- phyloseq::plot_bar(as_binary_otu_table(ecm_phy), x = "zone") 
ecm_phy_ggplot3 <- ggplot(ecm_phy_ggplot3$data, aes(fill= fct_reorder(as.factor(OTU_ID), Abundance), x=zone, y = Abundance))+
  geom_bar(stat="identity", position = "stack") +
  scale_fill_manual(values = funky(otu_nb_hala))+
  facet_grid(~year, scale="free")+
  labs(title="OTU Genus count (0;1) by distance from cut and by year")+
  xlab("position from SC")+
  guides(fill=guide_legend(title="Genus"))+
  theme(axis.text.x=element_text(color="coral4", size=8, angle=20))
ecm_phy_ggplot3
```

```{r boxplot}
# OTU richness 

hala_box1 <- ggplot( aes(x=name, y=value, fill=name)) +
    geom_boxplot() +
    scale_fill_viridis(discrete = TRUE, alpha=0.6) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    theme_ipsum() +
    theme(
      legend.position="none",
      plot.title = element_text(size=11)
    ) +
    ggtitle("A boxplot with jitter") +
    xlab("")

phyloseq::psmelt(ecm_phy) %>%
ggplot(data = ., aes(x = zone, y = Abundance)) +
  geom_boxplot(outlier.shape  = NA) +
  geom_jitter(aes(color = ), height = 0, width = 0.2) +
  labs(x = "Treatment", y = "Abundance")+
  facet_wrap(~ year)

phyloseq::psmelt(fanga_ecm) %>%
ggplot(data = ., aes(x = treatment, y = Abundance)) +
  geom_boxplot(outlier.shape  = NA) +
  geom_jitter(aes(color = ), height = 0, width = 0.2) +
  labs(x = "Treatment", y = "Abundance")+
  facet_wrap(~ year)

# abundance

hala_box2 <- boxplot_abundance(d= ecm_phy, x='zone', y='otu_table',
   line='subject')
```

## 3.2.2 multivariate analysis TY

```{r fruitbody type}
fruit_nb <- as.factor(ecm_phy@tax_table[,8]) %>% droplevels() %>% levels() %>% length() %>% print()
## Genus abundance by treatment
ecm_phy_fruit1 <- phyloseq::plot_bar(ecm_phy, x = "zone") 
ecm_phy_fruit1 <- ggplot(ecm_phy_fruit1$data, aes(fill= fct_reorder(as.factor(Fruitbody), Abundance), x = zone, y = Abundance))+
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = funky(fruit_nb))+
  facet_grid(~ year, scale="free")+
  labs(title="Genus abundance by distance from cut and by year")+
  xlab("Distance from cut")+
  guides(fill=guide_legend(title="Fruitbody"))+
  theme(axis.text.x=element_text(color="coral4", size=8, angle=20))
ecm_phy_fruit1

## % 
fruit_count <- ecm_phy_fruit1$data %>% group_by(Fruitbody, year, zone) %>%
  dplyr::summarise(sumcount = sum(Abundance, na.rm = TRUE)) %>% as.data.frame()

# chi-squared goodness-of-fit test
fruit12_chi <- chisq.test(x = fruit12, correct = F)
fruit18_chi <- chisq.test(x = fruit18, correct = F)
fruit12_chi
fruit18_chi

fruit12_chi$observed
round(fruit12_chi$expected, 0)
round(fruit12_chi$residuals, 1) # pearson residuals

fruit18_chi$observed
round(fruit18_chi$expected, 0)
round(fruit18_chi$residuals, 1) # pearson residuals

# GRAPHICAL
# 1. convert the data as a table
fruit12tab <- as.table(as.matrix(fruit12))
fruit18tab <- as.table(as.matrix(fruit18))
# 2. Graph
par(mfrow = c(1, 2))
balloonplot(t(fruit12tab), main ="", xlab ="", ylab="",
            label = FALSE, show.margins = FALSE)
balloonplot(t(fruit18tab), main ="", xlab ="", ylab="",
            label = FALSE, show.margins = FALSE)

# correlation plot
corrplot(fruit12_chi$residuals, is.cor = FALSE)
corrplot(fruit18_chi$residuals, is.cor = FALSE)
```

```{r OTU Genus composition}
# phylum
asc <- which(sum_res$data == "Ascomycota") %>% length()
bas <- which(sum_res$data == "Basidiomycota") %>% length()
pourc_asc <- (asc / (asc + bas))*100; pourc_asc
pourc_bas <- (bas / (asc + bas))*100; pourc_bas
# % of occurence among samples 

occurence_pourc_12 <- pourc_genusdf12b %>%
mutate(class = fct_reorder(Genus, pourc, .fun='max')) %>%
  ggplot( aes(x = reorder(Genus, pourc), y = pourc, fill = yr_treat_genus)) + 
  geom_bar(position = "stack", stat="identity", alpha = 1, width = 1, color = "black")+
  scale_fill_manual(values = coltreatment)+
  coord_flip()+
  labs(title = "2012")+
  scale_y_continuous(name = "Occurence of OTU for each sample type (in %)", breaks = seq(from = 0, to = 120, by = 20))+
  xlab("OTU Genus")+
  guides(fill = guide_legend(title = "Distance from cut"))+
  theme(axis.text.y = element_text(color="black", size = 4.5, angle=0), legend.position = c(0.85,0.3))
occurence_pourc_12

occurence_pourc_18 <- pourc_genusdf18b %>%
mutate(class = fct_reorder(Genus, pourc, .fun='max')) %>%
  ggplot( aes(x = reorder(Genus, pourc), y = pourc, fill = yr_treat_genus)) + 
  geom_bar(position = "stack", stat="identity", alpha = 1, width = 1, color = "black")+
  scale_fill_manual(values = coltreatment)+
  coord_flip()+
  labs(title = "2018")+
  scale_y_continuous(name = "Occurence of OTU for each sample type (in %)", breaks = seq(from = 0, to = 140, by = 20))+
  xlab("OTU Genus")+
  guides(fill = guide_legend(title = "Distance from cut"))+
  theme(axis.text.y = element_text(color="black", size = 4.5, angle=0), legend.position = c(0.85,0.3))
occurence_pourc_18

# number of singletons 
singleton_count12 <- length(which(colSums(osam12[, c(4:ncol(osam12))]) == 1))
total_count12 <-length(which(colSums(osam12[, c(4:ncol(osam12))]) >= 1))
pourc_sing12 <- singleton_count12/total_count12*100

singleton_count18 <-length(which(colSums(osam18[, c(4:ncol(osam18))]) == 1))
total_count18 <-length(which(colSums(osam18[, c(4:ncol(osam18))]) >= 1))
pourc_sing18 <- singleton_count18/total_count18*100

# print
singleton_count12; total_count12 # 2012: sing count; total count 
pourc_sing12 # 2012% singletons

singleton_count18; total_count18 # 2018: sing count; total count 
pourc_sing18 # 2018% singletons

## most abundant OTU genus :
### >25% occurrence in all samples

abun_otu12 <- which(colSums(pourc_genusdf12[, c(2:ncol(pourc_genusdf12))])/nrow(pourc_genusdf12) >= 25)
abun_otu12 
abun_otu18 <- which(colSums(pourc_genusdf18[, c(2:ncol(pourc_genusdf18))])/nrow(pourc_genusdf18) >= 25)
abun_otu18 

# most abundant and frequent genus:

# 2012: 
## Genus richness (and theirn nb of OTU)
colnames(ecm_12@tax_table)
tab_abund12 <- table(ecm_12@tax_table[, 5]) 
tab_abund12 <- tab_abund12[rev(order(tab_abund12))] 
tab_abund12

# 2018: 
## Genus richness (and theirn nb of OTU)
colnames(ecm_18@tax_table)
tab_abund18 <- table(ecm_18@tax_table[, 5]) 
tab_abund18 <- tab_abund18[rev(order(tab_abund18))] 

## Genus abundance (total read of OTU)
sum_by_level <- aggregate(Abundance ~ Genus+ year, data = ecm_phy_ggplot2$data, sum) %>% arrange(rev(desc(year)), desc(Abundance))
sum_by_level <- data.frame(sum_by_level)

## total nb of genus
genus_nb_hala <- as.factor(ecm_phy@tax_table[,5]) %>% droplevels() %>% levels() %>% length() %>% print()
## specific to 2012
spe2012 <- setdiff(ecm_12@tax_table[, 5], ecm_18@tax_table[, 5]); spe2012
## specific to 2018 
spe2018 <- setdiff(ecm_18@tax_table[, 5], ecm_12@tax_table[, 5]); spe2018

## specific to FOR 
spe_for <- setdiff(ecm_for@tax_table[, 5], union(ecm_edg@tax_table[, 5], ecm_cut@tax_table[, 5])); spe_for
## specific to CCFM
spe_edg <- setdiff(ecm_edg@tax_table[, 5], union(ecm_for@tax_table[, 5], ecm_cut@tax_table[, 5])); spe_edg
## specific to SC 
spe_cut <- setdiff(ecm_cut@tax_table[, 5], union(ecm_for@tax_table[, 5], ecm_edg@tax_table[, 5])); spe_cut

```

```{r multivariate representation}
# plotting OTUs instead of samples to see which OTU drive the samples in the ordination
# if possible, othewise, try CA

# jaccard
ecm_phy_ordi_mdsjacc <- ordinate(as_binary_otu_table(ecm_phy), method = "MDS", dist="jaccard")
ecm_phy_ordi_mdsjacc <- plot_ordination(ecm_phy, ecm_phy_ordi_mdsjacc, type = "samples", color="zone",label="sample_id")+
  geom_point(size=2)+
  facet_wrap(~year)+
  # color and elipse
  scale_color_manual(values = coltreatment_raref)+
  stat_ellipse(type="t", level=0.70)+
  # legend
  guides(color=guide_legend(title="Distance from cut"))+
  ggtitle(label = "MDS of samples grouped by distance from cut and by year", subtitle="presence-absence (Jaccard index)")
ecm_phy_ordi_mdsjacc

# bray
ecm_phy_ordi_mdsbray <- ordinate(ecm_phy, method = "MDS", dist="bray")
ecm_phy_ordi_mdsbray <- plot_ordination(ecm_phy, ecm_phy_ordi_mdsbray, type="samples", color="zone",label="sample_id")+
  geom_point(size=2)+
  # color and elipse
  scale_color_manual(values = coltreatment_raref)+
  stat_ellipse(type="t", level=0.70)+
  # legend
  guides(color=guide_legend(title="Distance from SC"))+
  ggtitle(label = "MDS of samples grouped by distance from cut and by year",subtitle = "number of sequence reads (bray-curtis index)")
ecm_phy_ordi_mdsbray

### BY YEAR INDEPENDANTLY 
# Remove empty samples 
which(colSums(ecm_12@otu_table) == 0)
which(colSums(ecm_18@otu_table) == 0) 

# s144: only one OTU, and specific to this sample 2012
# s72 : empty sample 2018
ecm_12_new = subset_samples(ecm_12, sample_id != "s144")
ecm_18_new = subset_samples(ecm_18, sample_id != "s72")

# 2012
jaccard12_ord <- ordinate(ecm_12, method = "MDS", dist="jaccard", binary = T)
jaccard12 <- plot_ordination(ecm_12, jaccard12_ord, type="samples", color="zone",label="sample_id")+
  geom_point(size=2)+
  # color and elipse
  scale_color_manual(values = coltreatment_raref)+
  stat_ellipse(type="t", level=0.70)+
  # legend
  guides(color=guide_legend(title="Distance from SC"))+
  ggtitle(label = "MDS of samples grouped by distance from cut for 2012",subtitle = "number of sequence reads (jaccard index)")
jaccard12

bray12_ord <- ordinate(ecm_12, method = "MDS", dist="bray")
bray12 <- plot_ordination(ecm_12, bray12_ord, type="samples", color="zone",label="sample_id")+
  geom_point(size=2)+
  # color and elipse
  scale_color_manual(values = coltreatment_raref)+
  stat_ellipse(type="t", level=0.70)+
  # legend
  guides(color=guide_legend(title="Distance from SC"))+
  ggtitle(label = "MDS of samples grouped by distance from cut for 2012",subtitle = "number of sequence reads (bray-curtis index)")
bray12

# 2018
jaccard18_ord <- ordinate(ecm_18, method = "MDS", dist="jaccard", binary = T)
jaccard18 <- plot_ordination(ecm_18, jaccard18_ord, type="samples", color="zone",label="sample_id")+
  geom_point(size=2)+
  # color and elipse
  scale_color_manual(values = coltreatment_raref)+
  stat_ellipse(type="t", level=0.70)+
  # legend
  guides(color=guide_legend(title="Distance from SC"))+
  ggtitle(label = "MDS of samples grouped by distance from cut for 2018",subtitle = "number of sequence reads (jaccard index)")
jaccard18

bray18_ord <- ordinate(ecm_18, method = "MDS", dist="bray")
bray18 <- plot_ordination(ecm_18, bray18_ord, type="samples", color="zone",label="sample_id")+
  geom_point(size=2)+
  scale_color_manual(values = coltreatment_raref)+
  stat_ellipse(type="t", level=0.70)+
  guides(color=guide_legend(title="Distance from SC"))+
  ggtitle(label = "MDS of samples grouped by distance from cut for 2018",subtitle = "number of sequence reads (bray-curtis index)")
bray18

## eigenvalues
ggplot()
barplot(jaccard12_ord$values$Eigenvalues)
barplot(bray12_ord$values$Eigenvalues)
barplot(jaccard18_ord$values$Eigenvalues)
barplot(bray18_ord$values$Eigenvalues)

### insight 
ecm_12@otu_table[, c("s36","s2","s113")] # just sharing Rhizopogon_mohelnensis c50
ecm_12@otu_table[, c("s6","s144","s89")] # sharing a lot of OTU with others samples
ecm_12@otu_table[, c("s142","s78")]
ecm_18@otu_table[, "s146"]
# do it with hellinger transformation (relative abundance)


# for specification_values

hala_ordi_specvalues_mdsjacc <- ordinate(as_binary_otu_table(ecm_phy), method = "MDS", dist="jaccard")
ecm_phy_ordi_mdsjacc <- plot_ordination(ecm_phy, hala_ordi_specvalues_mdsjacc, type="samples", color="specification_value",label="sample_id")+
  geom_point(size=2)+
  facet_wrap(~year)+
  # color and elipse
  scale_color_manual(values = c("N" = "lightblue",
                              "S" = "lightpink",
                              "UP" = "darkblue",
                              "LO" = "darkdarkred"))+
  stat_ellipse(type="t", level=0.70)+
  # legend
  guides(color=guide_legend(title="position from SC"))+
  ggtitle(label = "MDS of samples from Halasen_edge by year", subtitle="presence-absence (jaccard distance)")
ecm_phy_ordi_mdsjacc
```

```{r multivariate models}
perm <- adonis_pq(ecm_phy, "zone * year")
perm

perm_treatment <- adonis_pq(ecm_phy, "zone")
perm_treatment

perm_year <- adonis_pq(ecm_phy, "year")
perm_year

# separating years 
## best imo 
perm_12 <- adonis_pq(ecm_12, "zone")
perm_12 # SI
pw_perm12 <-pairwise.adonis2(t(ecm_12@otu_table) ~ zone, data = data.frame(ecm_12@sam_data))
pw_perm12 

perm_18 <- adonis_pq(ecm_18, "zone")
perm_18 # NS 

# corroborates what we are observing graphically 
```

```{r heatmap}
# distance heatmap (bray)
braydist = phyloseq::distance(ecm_phy, "bray")
dist_df = reshape2::melt(as.matrix(braydist))
colnames(dist_df) <- c("Samples1", "Samples2", "Bray_Curtis")

htmp_bray <- ggplot(dist_df, aes(Samples1, Samples2, fill = Bray_Curtis)) + 
  geom_tile()+
  theme(axis.text.x = element_text(angle = -90, vjust = 0.5, hjust = 0)) + 
  xlab("Samples") + ylab("Samples")
htmp_bray

# grouped heatmaps
## by treatment 
htmp_all <- plot_heatmap(phy_merged, method = "MDS",
             sample.label = "yr_treat", sample.order = order_level_mg,
             taxa.label = "OTU_ID", 
             low = "grey", high = "red", na.value = "white")
htmp_all

ht_count0 <- htmp_all$data %>% group_by(Genus, yr_treat) %>%
  dplyr::summarise(sumcount = sum(Abundance, na.rm = TRUE)) %>% as.data.frame()

ht_count <- ht_count0 %>%
  # convert state to factor and reverse order of levels
  mutate(Genus = factor(Genus, levels = rev(sort(unique(Genus))))) %>%
  # create a new variable from count
  mutate(countfactor = cut(sumcount, breaks = c(-1, 1, 5, 10, 50, 150, 450, max(sumcount, na.rm = T)),
                          labels=c("0", "0-1", "5-10", "10-50", "50-150", "150-450", ">450"))) %>%
  # change level order
  mutate(countfactor=factor(as.character(countfactor), levels=rev(levels(countfactor))))

### trying to creat multi xlabel
yr <- ifelse(grepl("2012", ht_count$yr_treat), "2012", "2018")
treat <- ifelse(grepl("CCFM", ht_count$yr_treat), "CCFM",
                         ifelse(grepl("SC", ht_count$yr_treat), "SC", "FOR")) # nested ifelse
ht_count <- cbind(ht_count, yr, treat)
### dont know.

## plotting
htmap_all <- ggplot(ht_count,  aes(x = factor(yr_treat, levels = order_level_mg), y = Genus, fill = countfactor)) + 
  geom_tile(colour = "white", size = 0.4)+
  scale_fill_manual(values = rev(c("grey90", RColorBrewer::brewer.pal(n = 5, name = "YlOrRd"))), na.value = "grey90")+
  theme_grey(base_size = 10)+
  guides(fill =guide_legend(title = str_wrap("Number of sequence reads (rarefied)", width = 20)))+
  labs(x = "Distance to cut and year", y = "Genus", title = "Occurence of OTU Genus by distance to cut and by year")+
  scale_x_discrete(labels = order_level_simpl)

htmap_all

# precise heatmap
htmap12 <- plot_heatmap(yr_12, method = "MDS", distance = "bray",
             sample.label = "zone", sample.order = order_level,
             taxa.label = "OTU_ID", 
             low = "grey", high = "red", na.value = "white")
htmap12

htmap18 <- plot_heatmap(yr_18, method = "MDS", distance = "bray",
             sample.label = "zone", sample.order = order_level,
             taxa.label = "OTU_ID",
             low = "grey", high = "red", na.value = "white")
htmap18

# heatmap pourc

htmap_genus <- ggplot(pourc_genusdf2,  aes(x = factor(yr_treat_genus, levels = order_level_mg), y = Genus, fill = pourc)) + 
  geom_tile(colour = "white", size = 0.4)
htmap_genus
```

```{r circle graph}
circle_pq(ecm_phy, "zone", taxa = "Genus", add_nb_seq = F,
          grid_col = "darkgrey",
          row_col = coltreatment)
```

# *10 - H2) HALASEN_5P STUDY: thinning effect - randomly distributed retention trees*

## 4.1 design and sampling

```{r overview }
colnames(hala5_sam)
summary(hala5_sam[,c(2,5:8)])
summary(hala5_sam$plot)
table(hala5_sam$plot,hala5_sam$treatment)
print(hala5_ecm@sam_data)

# nb genus
colnames(hala5_ecm@tax_table)
as.factor(hala5_ecm@tax_table[,5]) %>% droplevels() %>% levels()
genus_nb_hala5 <- as.factor(hala5_ecm@tax_table[,5]) %>% droplevels() %>% levels() %>% length() %>% print() # 18 genus
```

## 4.2.1 bivariate analysis

```{r rarecurve}
hala5_rrcurve0 <- phyloseq.extended::ggrare(hala5_grp, step = 10, color = "treatment",plot=F, se=T)+
  labs(x = "Group size",
       y =  "OTU richness",
       title = "rarefaction curves by treatment",
       color = "Treatment")+
  guides(fill = F)+ # Suppressing the filling (standard error) legend 
  scale_color_manual(values = c("forest" = "darkgreen",
                                "thinned" = "orange",
                                "clearcut" = "darkred"))+
  scale_fill_manual(values = c("forest" = "darkgreen",
                                "thinned" = "orange",
                                "clearcut" = "darkred"))
hala5_rrcurve0
```

```{r circle graph: genus distribution among treatments}
circle_pq(physeq = hala5_ecm, 
                              fact = "treatment", 
                              taxa = "Genus", 
                              min_prop_mod=0.001,
                              min_prop_tax=0.01,
                              row_col = c("clearcut" = "darkred",
                                          "forest" ="darkgreen",
                                          "thinned" = "orange"),
                              grid_col= 1,
                              add_nb_seq = T) # distribution circle sample - taxonomy 
```

```{r genus distribution}
genus_nb_hala5 <- as.factor(hala5_ecm@tax_table[,5]) %>% droplevels() %>% levels() %>% length()
hala5_ecm_ggplot1 <- phyloseq::plot_bar(as_binary_otu_table(hala5_ecm), x = "treatment") 
hala5_ecm_ggplot1 <- ggplot(hala5_ecm_ggplot1$data, aes(fill= fct_reorder(as.factor(Genus), Abundance), x=treatment, y = Abundance))+
  geom_bar(stat="identity", position = "stack") +
  scale_fill_manual(values = cols25(genus_nb_hala5))+
  facet_grid(~year, scale="free")+
  labs(title="OTU Genus count (0;1) by cutting treatment",subtitle = "Halasen_P5")+
  guides(fill=guide_legend(title="Genus"))
hala5_ecm_ggplot1

hala5_ecm_ggplot2 <- phyloseq::plot_bar(hala5_ecm, x = "treatment") 
hala5_ecm_ggplot2 <- ggplot(hala5_ecm_ggplot2$data, aes(fill= fct_reorder(as.factor(Genus), Abundance), x=treatment, y = Abundance))+
  geom_bar(stat="identity", position = "stack") +
  scale_fill_manual(values = cols25(genus_nb_hala5))+
  facet_grid(~year, scale="free")+
  labs(title="OTU Genus abundance (nb of reads) by cutting treatment",subtitle = "Halasen_P5")+
  guides(fill=guide_legend(title="Genus"))
hala5_ecm_ggplot2
```

```{r OTUs distribution}
otu_nb_hala5 <- as.factor(hala5_ecm@tax_table[,6]) %>% droplevels() %>% levels() %>% length()
hala5_ecm_ggplot3 <- phyloseq::plot_bar(as_binary_otu_table(hala5_ecm), x = "treatment")
hala5_ecm_ggplot3 <- ggplot(hala5_ecm_ggplot3$data, aes(fill= fct_reorder(as.factor(OTU_ID), Abundance), x=treatment, y = Abundance))+
  geom_bar(stat="identity", position = "stack") +
  scale_fill_manual(values = funky(otu_nb_hala5))+
  facet_grid(~year, scale="free")+
  labs(title="OTU Genus count (0;1) by cutting treatment",subtitle = "Halasen_P5")+
  guides(fill=guide_legend(title="Genus"))
hala5_ecm_ggplot3

hala5_ecm_ggplot4 <- phyloseq::plot_bar(hala5_ecm, x = "treatment") 
hala5_ecm_ggplot4 <- ggplot(hala5_ecm_ggplot4$data, aes(fill= fct_reorder(as.factor(OTU_ID), Abundance), x=treatment, y = Abundance))+
  geom_bar(stat="identity", position = "stack") +
  scale_fill_manual(values = funky(otu_nb_hala5))+
  facet_grid(~year, scale="free")+
  labs(title="OTU abundance (nb of reads) by cutting treatment",subtitle = "Halasen_P5")+
  guides(fill=guide_legend(title="Genus"))
hala5_ecm_ggplot4
```

```{r hill OTUs}
hala5_hill_treatment <- hill_pq(physeq = hala5_ecm_t, variable = "treatment", color_fac = NA, letters = F) 
hala5_hill0 <- hala5_hill_treatment[[1]]+ 
  labs(title="Hill 0 : apha diversity of ", subtitle="Halasen5")+
  coord_cartesian(xlim=c(0,16))+
  scale_x_continuous(name="Hill_0", breaks=seq(0,16,by=2))+
  scale_y_continuous(name="treatment", breaks=c(1,2,3))+
  scale_color_manual(values = c("forest" = "darkgreen",
                                "thinned" = "orange",
                                "clearcut" = "darkred"))+
  theme(axis.text.x=element_text(color="coral4", size=8, angle=0))
hala5_hill0

hala5_hill1 <- hala5_hill_treatment[[2]]+ 
  labs(title="Hill 1 : treatment", subtitle="Halasen5")+
  coord_cartesian(xlim=c(0,12))+
  scale_x_continuous(name="Hill_1", breaks=seq(0,12,by=2))+
  scale_y_continuous(name="treatment", breaks=c(1,2,3))+
  scale_color_manual(values = c("forest" = "darkgreen",
                                "thinned" = "orange",
                                "clearcut" = "darkred"))+
  theme(axis.text.x=element_text(color="coral4", size=8, angle=0))

hala5_hill2 <- hala5_hill_treatment[[3]]+ 
  labs(title="Hill 2 : treatment", subtitle="Halasen5")+
  coord_cartesian(xlim=c(0,10))+
  scale_x_continuous(name="Hill_2", breaks=seq(0,10,by=2))+
  scale_y_continuous(name="treatment", breaks=c(1,2,3))+
  scale_color_manual(values = c("forest" = "darkgreen",
                                "thinned" = "orange",
                                "clearcut" = "darkred"))+
  theme(axis.text.x=element_text(color="coral4", size=8, angle=0))

hala5_hill_treatment[[4]]
hala5_hill_treatment[[4]] # TukeyHSD test
```

## 4.2.2 multivariate analysis

```{r ordination}
hala5_ecm_ordi_mdsjacc  <- ordinate(hala5_ecm, method = "MDS", dist="bray")
hala5_ecm_ordi_mdsjacc  <- plot_ordination(hala5_ecm, hala5_ecm_ordi_mdsjacc, type="samples", color="treatment",label="sample_id")+
  geom_point(size=2)+
  # color and elipse
  scale_color_manual(values = c("clearcut" = "darkred",
                                "thinned" = "orange",
                                "forest" = "darkgreen"))+
  stat_ellipse(type="t", level=0.70)+
  # legend
  guides(color=guide_legend(title="treatment"))+
  ggtitle(label = "MDS of samples from Halasen_5P",subtitle = "abundance")
hala5_ecm_ordi_mdsjacc

hala5_ecm_ordi_mdsbray  <- ordinate(as_binary_otu_table(hala5_ecm), method = "MDS", dist="jaccard")
hala5_ecm_ordi_mdsbray  <-plot_ordination(hala5_ecm, hala5_ecm_ordi_mdsbray , type="samples", color="treatment",label="sample_id")+
  geom_point(size=2)+
  # color and elipse
    scale_color_manual(values = c("clearcut" = "darkred",
                                "thinned" = "orange",
                                "forest" = "darkgreen"))+
  stat_ellipse(type="t", level=0.70)+
  # legend
  guides(color=guide_legend(title="treatment"))+
  ggtitle(label = "MDS of samples from Halasen_5P",subtitle = "abundance")

hala5_ecm_ordi1 <- ordinate(as_binary_otu_table(hala5_ecm), method = "NMDS", dist="jaccard",k=20)
plot_ordination(hala5_ecm, hala5_ecm_ordi1, type="samples", color="treatment",label="sample_id")+
  geom_point(size=2)+
  # color and elipse
    scale_color_manual(values = c("clearcut" = "darkred",
                                "thinned" = "orange",
                                "forest" = "darkgreen"))+
  stat_ellipse(type="t", level=0.70)+
  # legend
  guides(color=guide_legend(title="treatment"))+
  ggtitle(label = "MDS of samples from Halasen_5P",subtitle = "abundance")

hala5_ecm_ordi_phyl1 <- ordinate(hala5_ecm, method = "MDS", dist="bray")
hala5_ecm_ordi_phyl1 <- plot_ordination(hala5_ecm, hala5_ecm_ordi_phyl1, type="taxa", color="Order", shape="Phylum")+
  geom_point(size=2)+
  # color and elipse
  scale_color_manual(values =cols25(10))+
  stat_ellipse(type="t", level=0.70)+
  # legend
  guides(color=guide_legend(title="position from SC"))+
  ggtitle(label = "MDS of samples from Halasen_edge by year",subtitle = "abundance (bray-curtis distance)")
hala5_ecm_ordi_phyl1
```

# *11 - F1) FANGAMON STUDY: thinning effect - randomly distributed retention trees*

## 5.1 design and sampling

```{r overview }
colnames(fanga_sam)
summary(fanga_sam[,c(2,5:8)])
summary(fanga_sam$plot)
table(fanga_sam$plot,fanga_sam$treatment)

# nb genus
colnames(fanga_ecm@tax_table)
as.factor(fanga_ecm@tax_table[,5]) %>% droplevels() %>% levels()
genus_nb_fanga <- as.factor(fanga_ecm@tax_table[,5]) %>% droplevels() %>% levels() %>% length()
genus_nb_fanga # 15 genus

```

## 5.2.1 bivariate analysis

```{r rarecurve}
fanga_rrcurve0 <- phyloseq.extended::ggrare(fanga_grp_12, step = 10, color = "treatment",plot=F, se=T)+ facet_wrap(~year)+
  labs(x = "Group size",
       y =  "OTU richness",
       title = "rarefaction curves by treatment",
       color = "Treatment")+
  guides(fill = F)+ # Suppressing the filling (standard error) legend 
  scale_color_manual(values = c("clearcut" = "darkred",
                                "forest" = "darkgreen",
                                "thinned_burn" = "grey4",
                                "thinned_no_burn" = "orange"))+
  scale_fill_manual(values = c("clearcut" = "darkred",
                                "forest" = "darkgreen",
                                "thinned_burn" = "grey4",
                                "thinned_no_burn" = "orange"))
fanga_rrcurve0

fanga_rrcurve1 <- phyloseq.extended::ggrare(fanga_grp_18, step = 10, color = "treatment",plot=F, se=T)+ facet_wrap(~year)+
  labs(x = "Group size",
       y =  "OTU richness",
       title = "rarefaction curves by treatment",
       color = "Treatment")+
  guides(fill = F)+ # Suppressing the filling (standard error) legend 
  scale_color_manual(values = c("clearcut" = "darkred",
                                "forest" = "darkgreen",
                                "thinned_burn" = "grey4",
                                "thinned_no_burn" = "orange"))+
  scale_fill_manual(values = c("clearcut" = "darkred",
                                "forest" = "darkgreen",
                                "thinned_burn" = "grey4",
                                "thinned_no_burn" = "orange"))
fanga_rrcurve1
```

```{r circle graph: genus distribution among treatments}
circle_fanga_ecm <- circle_pq(physeq = fanga_ecm, 
                              fact = "treatment", 
                              taxa = "Genus", 
                              min_prop_mod=0.001,
                              min_prop_tax=0.02,
                              row_col = c("clearcut" = "darkred",
                                "forest" = "darkgreen",
                                "thinned_burn" = "grey4",
                                "thinned_no_burn" = "orange"),
                              grid_col= 1,
                              add_nb_seq = T)
```

```{r genus distribution}
genus_nb_fanga <- as.factor(fanga_ecm@tax_table[,5]) %>% droplevels() %>% levels() %>% length()
fanga_ecm_ggplot1 <- phyloseq::plot_bar(as_binary_otu_table(fanga_ecm), x = "treatment") 
fanga_ecm_ggplot1 <- ggplot(fanga_ecm_ggplot1$data, aes(fill= fct_reorder(as.factor(Genus), Abundance), x=treatment, y = Abundance))+
  geom_bar(stat="identity", position = "stack") +
  facet_grid(~year, scale="free")+
  scale_fill_manual(values = cols25(genus_nb_fanga))+
  facet_grid(~year, scale="free")+
  labs(title="OTU Genus count (0;1) by cutting treatment",subtitle = "Fangamon")+
  guides(fill=guide_legend(title="Genus"))
fanga_ecm_ggplot1

fanga_ecm_ggplot2 <- phyloseq::plot_bar(fanga_ecm, x = "treatment") 
fanga_ecm_ggplot2 <- ggplot(fanga_ecm_ggplot2$data, aes(fill= fct_reorder(as.factor(Genus), Abundance), x=treatment, y = Abundance))+
  geom_bar(stat="identity", position = "stack") +
  facet_grid(~year, scale="free")+
  scale_fill_manual(values = cols25(genus_nb_fanga))+
  facet_grid(~year, scale="free")+
  labs(title="OTU Genus abundance (nb of reads) by cutting treatment",subtitle = "Fangamon")+
  guides(fill=guide_legend(title="Genus"))
fanga_ecm_ggplot2
# position = fill -> relative abundance 
```

```{r OTU distribution}
otu_nb_fanga <- as.factor(fanga_ecm@tax_table[,6]) %>% droplevels() %>% levels() %>% length()
fanga_ecm_ggplot3 <- phyloseq::plot_bar(as_binary_otu_table(fanga_ecm), x = "treatment") 
fanga_ecm_ggplot3 <- ggplot(fanga_ecm_ggplot3$data, aes(fill= fct_reorder(as.factor(OTU_ID), Abundance), x=treatment, y = Abundance))+
  geom_bar(stat="identity", position = "stack") +
  facet_grid(~year, scale="free")+
  scale_fill_manual(values = funky(otu_nb_fanga))+
  facet_grid(~year, scale="free")+
  labs(title="OTU Genus count (0;1) by cutting treatment",subtitle = "Fangamon")+
  guides(fill=guide_legend(title="Genus"))
fanga_ecm_ggplot3

fanga_ecm_ggplot4 <- phyloseq::plot_bar(fanga_ecm, x = "treatment") 
fanga_ecm_ggplot4 <- ggplot(fanga_ecm_ggplot4$data, aes(fill= fct_reorder(as.factor(OTU_ID), Abundance), x=treatment, y = Abundance))+
  geom_bar(stat="identity", position = "stack") +
  facet_grid(~year, scale="free")+
  scale_fill_manual(values = funky(otu_nb_fanga))+
  facet_grid(~year, scale="free")+
  labs(title="OTU Genus abundance (nb of reads) by cutting treatment",subtitle = "Fangamon")+
  guides(fill=guide_legend(title="Genus"))
fanga_ecm_ggplot4
# position = fill -> relative abundance 
```

```{r hill fanga}
fanga_hill_treatment <- hill_pq(physeq = fanga_ecm_t,
                                variable = "treatment", 
                                color_fac = NA, 
                                letters = F) 

fanga_hill0 <- fanga_hill_treatment[[1]]+ 
  labs(title="Hill 0 : treatment", subtitle="Fangamon")+
  coord_cartesian(xlim=c(0,16))+
  scale_x_continuous(name="Hill_0", breaks=seq(0,16,by=2))+
  scale_y_continuous(name="treatment", breaks=c(1,2,3))+
  scale_color_manual(values = c("clearcut" = "darkred",
                                "forest" = "darkgreen",
                                "thinned_burn" = "grey4",
                                "thinned_no_burn" = "orange"))+
  theme(axis.text.x=element_text(color="coral4", size=8, angle=0))
fanga_hill0

fanga_hill1 <- fanga_hill_treatment[[2]]+ 
  labs(title="Hill 1 : treatment", subtitle="Fangamon")+
  coord_cartesian(xlim=c(0,12))+
  scale_x_continuous(name="Hill_1", breaks=seq(0,12,by=2))+
  scale_y_continuous(name="treatment", breaks=c(1,2,3))+
  scale_color_manual(values = c("clearcut" = "darkred",
                                "forest" = "darkgreen",
                                "thinned_burn" = "grey4",
                                "thinned_no_burn" = "orange"))+
  theme(axis.text.x=element_text(color="coral4", size=8, angle=0))

fanga_hill2 <- fanga_hill_treatment[[3]]+ 
  labs(title="Hill 2 : treatment", subtitle="Fangamon")+
  coord_cartesian(xlim=c(0,10))+
  scale_x_continuous(name="Hill_2", breaks=seq(0,10,by=2))+
  scale_y_continuous(name="treatment", breaks=c(1,2,3))+  
  scale_color_manual(values = c("clearcut" = "darkred",
                                "forest" = "darkgreen",
                                "thinned_burn" = "grey4",
                                "thinned_no_burn" = "orange"))+
  theme(axis.text.x=element_text(color="coral4", size=8, angle=0))

fanga_hill_treatment[[4]]
fanga_hill_treatment[[4]] # TukeyHSD test
```

```{r trying understand difference between boolean and relative abundance}
fanga_ecm_only_SC <- subset_samples(fanga_ecm, fanga_ecm@sam_data$treatment=="clearcut") %>% clean_physeq(remove_empty_samples = T,remove_empty_taxa = T)

print(fanga_ecm_only_SC@otu_table)
which(fanga_ecm_only_SC@otu_table>200,arr.ind = T) %>% rownames() 
fanga_ecm_only_SC@tax_table[c("c10", "c21"),6] # 2 super abundant species 
```

```{r maybe a contamination for a sample?}
fanga_ecm_only_SC_2012 <- subset_samples(fanga_ecm, fanga_ecm@sam_data$year=="2012" & fanga_ecm@sam_data$treatment=="clearcut") %>% clean_physeq(remove_empty_samples = T,remove_empty_taxa = T)
print(fanga_ecm_only_SC_2012@sam_data)
print(fanga_ecm_only_SC_2012@otu_table) # c10 in both samples, c21 in another different one... no specific contamination it seems

fanga_ecm_excl_samples <- subset_samples(fanga_ecm, !fanga_ecm@sam_data$sample_id %in% c("s152","s45")) %>% clean_physeq(remove_empty_samples = T,remove_empty_taxa = T)

fanga_ecm_ggplot3 <- phyloseq::plot_bar(as_binary_otu_table(fanga_ecm_excl_samples), x = "treatment") 
ggplot(fanga_ecm_ggplot3$data, aes(fill= fct_reorder(as.factor(Genus), Abundance), x=treatment, y = Abundance))+
  geom_bar(stat="identity", position = "stack") +
  facet_grid(~year, scale="free")+
  scale_fill_manual(values = cols25(genus_nb_fanga))+
  facet_grid(~year, scale="free")+
  labs(title="OTU Genus count (0;1) by cutting treatment",subtitle = "Fangamon- exlcuded s152,s45 samples")+
  guides(fill=guide_legend(title="Genus"))

fanga_ecm_ggplot2 <- phyloseq::plot_bar(fanga_ecm_excl_samples, x = "treatment") 
ggplot(fanga_ecm_ggplot2$data, aes(fill= fct_reorder(as.factor(Genus), Abundance), x=treatment, y = Abundance))+
  geom_bar(stat="identity", position = "stack") +
  facet_grid(~year, scale="free")+
  labs(title="", subtitle="")  
```

## 5.2.2 multivariate analysis

```{r ordination}
fanga_ecm_ordi_mdsjacc <- ordinate(as_binary_otu_table(fanga_ecm), method = "MDS", dist="jaccard")
fanga_ecm_ordi_mdsjacc <- plot_ordination(fanga_ecm, fanga_ecm_ordi_mdsjacc, type="samples", color="treatment",label="sample_id")+
  geom_point(size=2)+
  facet_wrap(~year)+
  # color and elipse
  scale_color_manual(values = c("clearcut" = "darkred",
                              "forest" = "darkgreen",
                              "thinned_burn" = "grey4",
                              "thinned_no_burn" = "orange"))+
  stat_ellipse(type="t", level=0.70)+
  # legend
  guides(color=guide_legend(title="position from SC"))+
  ggtitle(label = "MDS of samples from fangamon by year", subtitle="presence-absence (jaccard distance)")
fanga_ecm_ordi_mdsjacc

fanga_ecm_ordi_mdsbray <- ordinate(fanga_ecm, method = "MDS", dist="bray")
fanga_ecm_ordi_mdsbray <- plot_ordination(fanga_ecm, fanga_ecm_ordi_mdsbray, type="samples", color="treatment",label="sample_id")+
  geom_point(size=2)+
  facet_wrap(~year)+
  # color and elipse
  scale_color_manual(values = c("clearcut" = "darkred",
                              "forest" = "darkgreen",
                              "thinned_burn" = "grey4",
                              "thinned_no_burn" = "orange"))+
  stat_ellipse(type="t", level=0.70)+
  # legend
  guides(color=guide_legend(title="position from SC"))+
  ggtitle(label = "MDS of samples from fangamon by year",subtitle = "abundance (bray-curtis distance)")
```

# *12 - EXPORT FILES*

```{r manuscript}
getwd()
setwd("H:/REDACTION/FIGURES/FIGURES_R")

## export tables as csv
write.csv(varsum, file = 'varsum.csv')

#### EXPORT 07-09-23
plot_list <- list(bxplt_rich, bxplt_relat, bray_fanga12, bray_fanga18, bray_hala12, bray_hala18, bray_hala5p, htmap_fanga, htmap_hala, htmap_hala5p)

output_dir <- "C:/Users/Utilisateur/OneDrive/STAGE-2023_SUEDE/DATA/PAPER_RESULTS/"

for (i in 1:length(plot_list)) {
  plot_file <- paste0(output_dir, "plot_", i, ".jpg")
  ggsave(plot_list[[i]], filename = plot_file, width = 8, height = 6)
}
```
